---
title: "Stochastic simulation"
subtitle: "PCB 6455 / BSC 4933: Biometry &nbsp;&nbsp;|&nbsp;&nbsp; Lecture 13"
author: "Dr. Brian Maitner"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    code-line-numbers: false
    fig-width: 7
    fig-height: 4.0
    fig-align: center
    html-math-method: mathml
    include-in-header:
      text: |
        <style>
        .type-it { border-left: 6px solid #2a7ab0; padding-left: 0.4em; }
        .type-it::before, .type-badge {
          content: "Type and run"; display: inline-block;
          background: #2a7ab0; color: white; font-weight: bold;
          font-size: 0.5em; line-height: 1.4; padding: 0 0.6em;
          border-radius: 4px; vertical-align: middle;
        }
        .type-badge { font-size: 0.7em; }
        </style>
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(emdbook)
data(ReedfrogFuncresp)
frogs <- ReedfrogFuncresp
set.seed(101)
```

## Today's agenda

- What a simulation is, and why you would run one
- Building a simulation: a curve, plus noise
- Groups, and how to simulate them
- The midterm

. . .

[Type and run]{.type-badge} on a code block: type it into your script and run it.
You will need it later. Code without the label is there to watch.

::: {.notes}
Chapter 5. This is the first lecture where they produce data rather than
describe it.

The through line: a simulation is chapter 3 plus chapter 4. A deterministic
function for the pattern, a distribution for the scatter. They have both halves
already.
:::

## What is a simulation?

**Experimentation with the help of a model.**

Choose a model, choose parameters, generate data, see what happens.

. . .

You already have both pieces:

- **Chapter 3:** a function for the expected value
- **Chapter 4:** a distribution for the noise around it

::: {.notes}
Worth saying plainly that nothing new is being introduced today, only combined.
They wrote Ricker and Michaelis-Menten functions in chapter 3 and met rnorm,
rbinom and rnbinom in chapter 4.

The one new idea is direction. Until now data came first and the model was
fitted to it. Today the model comes first and the data come out of it.
:::

## Why simulate?

- **Study design.** How many samples do I need to detect this effect?
- **Model checking.** Does my model produce data that look like my data?
- **Method checking.** Does this analysis recover the answer I put in?
- **Theory.** What happens if I add this mechanism?

. . .

The second and third are why this is not optional: it is how you find out
whether your analysis works **before** you trust it.

::: {.notes}
Study design is Thursday's lecture, power analysis, and part 3 of the midterm.

Method checking is the one worth dwelling on. If you simulate data with a known
slope of 1.5 and your analysis says 0.4, the analysis is wrong, and you found
out without needing a single real measurement. Real data cannot do that for you,
because you never know the true answer.

Bolker's framing in 5.1: simulation is how you make sure the pieces work
together before you commit to them.
:::

## The recipe

1. Pick a model, the deterministic part
2. Pick parameters
3. Add noise from a distribution
4. Look at the output
5. Change something and repeat

::: {.notes}
Five steps, and the rest of the lecture is three passes through them with
different pieces swapped in: a line with normal noise, a curve with negative
binomial noise, then groups.

Step 5 is where the learning happens, and it is also the step students skip.
:::

## Start with the deterministic part {.smaller}

::: {.type-it}
```{r}
#| fig-height: 2.4
a <- 2
b <- 1
x <- 1:20

y_det <- a + b * x

plot(x = x, y = y_det)
```
:::

A line: $y = a + bx$. No randomness yet.

::: {.notes}
Deliberately the simplest possible model, so the attention is on the structure
rather than on the function.

Note that y_det is a vector of 20 expected values, one per x. The noise gets
added to each of them in a moment.
:::

## Now add noise

::: {.type-it}
```{r}
#| fig-height: 2.8
y_stoch <- rnorm(n = 20, mean = y_det, sd = 2)

plot(x = x, y = y_stoch)
```
:::

`mean = y_det` is the trick: **a different mean for every point.**

::: {.notes}
This is the step that makes it a statistical model, and the vectorized mean is
the part worth pointing at. rnorm is given 20 means and returns 20 draws, one
around each.

Say it as a sentence: each observation is its expected value plus noise. That
sentence is every model in the rest of the course.
:::

## Run it again

::: {.type-it}
```{r}
#| eval: false
y_stoch <- rnorm(n = 20, mean = y_det, sd = 2)
plot(x = x, y = y_stoch)
```
:::

```{r}
#| echo: false
#| fig-height: 3.2
par(mfrow = c(1, 4), mar = c(3.2, 3.2, 1.6, 0.6), mgp = c(1.9, 0.6, 0))
for (i in 1:4) {
  ys <- rnorm(n = 20, mean = y_det, sd = 2)
  plot(x, ys, pch = 16, col = "grey30", ylim = c(0, 28),
       xlab = "x", ylab = "y", main = paste("run", i))
  abline(a = 2, b = 1, col = "red", lwd = 2)
}
```

Same model, same parameters, four different datasets.

::: {.notes}
Have them run it several times themselves before showing this.

The point is the one students find genuinely surprising: the data are not the
model. Each dataset is one draw from a process, and the red line, the truth, is
the same in all four panels.

This is also the honest version of "my result did not replicate". Some of that
spread is just this.
:::

## When the noise grows with the mean

::: {.type-it}
```{r}
#| fig-height: 2.7
sd_multiplier <- 0.3

y_stoch <- rnorm(n = 20, mean = y_det, sd = y_det * sd_multiplier)

plot(x = x, y = y_stoch)
points(x, y_det, col = "blue")
```
:::

Bigger things vary more. Very common in biology.

::: {.notes}
Body size, abundance, biomass: in all of them the spread grows with the mean,
which is why a constant sd is often wrong.

Connect to Lecture 12: this is the multiplicative idea again. Constant
proportional variation is what makes data lognormal.

The blue points are the deterministic expectation, so the fan shape is visible
against it.
:::

## Your turn: break it {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
a <- 2
b <- 1
x <- 1:20
y_det <- a + b * x

y_stoch <- rnorm(n = 20, mean = y_det, sd = 2)
plot(x = x, y = y_stoch)
abline(a = a, b = b, col = "red")

# now try
# sd = 0.1    sd = 20    b = 0    b = -1
```
:::

How big does `sd` have to be before the line is invisible in the data?

::: {.notes}
Six minutes. The last question has no single right answer and that is the point:
whether a pattern is visible depends on the ratio of signal to noise, not on
either one alone.

With sd = 20 and this slope, the line is hopeless with 20 points. That is
exactly the question power analysis answers on Thursday, so this exercise is the
setup for it.

b = 0 is worth doing too: it shows what "no effect" data look like, which is the
thing they will be testing against all semester.
:::

## A curve instead of a line

Bolker's example: a hyperbolic function with **negative binomial** noise.

$$y = \frac{a\,b}{b + x}$$

```{r}
#| echo: false
#| fig-height: 2.6
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
a2 <- 20; b2 <- 1
xx <- runif(50, min = 0, max = 5)
plot(xx, a2 * b2 / (b2 + xx), pch = 16, col = "grey30",
     xlab = "x", ylab = "expected y")
```

Why would a normal error term be wrong here?

::: {.notes}
Section 5.2 in the book. Two things are new: the deterministic part is a curve,
and the noise is a count distribution.

Answer to the question on the slide: y is a count, so it cannot be negative and
it cannot be fractional, and near the right side the expected value is small
enough that a normal would predict negative counts. Counts need a count
distribution.

Note runif for the x values: the predictor itself is random here, which is what
an observational study looks like. Worth asking how that differs from x <- 0:5,
which is a designed experiment with fixed levels.
:::

## Adding negative binomial noise {.smaller}

::: {.type-it}
```{r}
#| fig-height: 2.3
a <- 20
b <- 1
k <- 5
x <- runif(50, min = 0, max = 5)

y_det <- a * b / (b + x)
y <- rnbinom(n = 50, mu = y_det, size = k)

plot(x = x, y = y)
```
:::

::: {.notes}
Run it several times. The scatter is wider at the left, where the mean is
larger, which is what a negative binomial does automatically: its variance grows
with its mean.

k is the clumping parameter from Lecture 12, the one they estimated for
cardinals as 1.8. Small k, more scatter.

Note the model written correctly as a * b / (b + x). With b = 1 that is the same
as a / (b + x), which is why the old version of this slide looked fine, but the
two differ as soon as b is not 1, which happens on the group slides shortly.
:::

## Your turn: what does each parameter do? {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
a <- 20
b <- 1
k <- 5
x <- runif(50, min = 0, max = 5)

y_det <- a * b / (b + x)
y <- rnbinom(n = 50, mu = y_det, size = k)

plot(x = x, y = y, ylim = c(0, 60))
points(x, y_det, col = "blue", pch = 16)
```
:::

Try `k = 0.5` and `k = 100`. Then `a = 50`. Then `b = 3`.

::: {.notes}
Seven minutes.

What to listen for: a sets the starting height, b controls how fast the curve
drops, and k controls only the scatter, not the expectation. That last one is
the distinction that matters: a and b are the biology, k is the noise.

k = 100 looks almost Poisson, k = 0.5 is wild. Same expected curve underneath
both, which is the blue points.
:::

## Groups {.center}

## Different species, different parameters

Two species, 25 observations each. They differ in `a` and `b`, but share the
same noise.

::: {.type-it}
```{r}
g <- factor(rep(1:2, each = 25))

a <- c(20, 10)
b <- c(1, 2)
```
:::

`a` and `b` are now **vectors**, one value per group.

::: {.notes}
This is the structural idea from the end of Lecture 12, arriving as something
they build rather than something to watch out for.

factor() matters: g is a grouping label, not the number 1 or 2. R will use it to
pick parameters and to pick colors.
:::

## Indexing by group {.smaller}

::: {.type-it}
```{r}
#| fig-height: 2.2
x <- runif(50, min = 0, max = 5)

y_det <- a[g] * b[g] / (b[g] + x)

plot(x = x, y = y_det, col = g, pch = 16)
```
:::

`a[g]` reads "the `a` for whichever group this observation belongs to".

::: {.notes}
a[g] is the piece of R syntax worth slowing down on. g is a factor of 1s and 2s,
so a[g] is a 50-long vector made of a[1] and a[2] in the right places.

Have them try a[1] and a[2] on their own to see the pieces.

col = g colors the points by group, because R converts the factor to 1 and 2 and
those are the first two colors in the palette. pch = g works the same way for
symbols.
:::

## Now add the noise

::: {.type-it}
```{r}
#| fig-height: 2.8
k <- 5

y <- rnbinom(n = 50, mu = y_det, size = k)

plot(x = x, y = y, col = g, pch = as.numeric(g))
legend("topright", bty = "n", col = 1:2, pch = 1:2,
       legend = c("species 1", "species 2"))
```
:::

::: {.notes}
Plotting it badly first is worth doing: without col and pch the two species are
indistinguishable, which is the honest version of the question "could I even
tell these apart in real data?".

The legend is the Lecture 8 habit arriving in base graphics.

Ask the room: do the two species look different here? With k = 5 and this much
overlap it is genuinely arguable, which leads into the next slide.
:::

## Your turn: can you tell them apart? {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
a <- c(20, 10)      # change these
b <- c(1, 2)
k <- 5

y_det <- a[g] * b[g] / (b[g] + x)
y <- rnbinom(n = 50, mu = y_det, size = k)
plot(x = x, y = y, col = g, pch = as.numeric(g))
```
:::

Find parameters where the two species look **clearly different**.
Then find parameters where they look **identical**.
Then find parameters where they differ only at **low x**.

::: {.notes}
Eight minutes, and the best exercise in the lecture. It is a real research
question in disguise: how different do two groups have to be before you could
detect it with 25 observations each?

The third case, differing only at low x, is the interesting one. It is common in
real data and it is why where you sample matters as much as how much you sample.
Anyone who finds it has learned something about study design that no amount of
reading would teach.

Raising k makes everything easier to see, which is the same lesson as the sd
exercise earlier.
:::

## Does my model make data like my data? {.center}

## The tadpoles, again

```{r}
#| echo: false
#| fig-height: 3.2
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
plot(frogs$Initial, frogs$Killed, pch = 16, cex = 1.5, col = "black",
     xlab = "tadpoles in the tank", ylab = "number killed",
     xlim = c(0, 100), ylim = c(0, 40))
curve(60 * x / (100 + x), add = TRUE, col = "red", lwd = 2)
```

Lecture 10: the curve. Lecture 12: killed **out of** Initial, so binomial.

::: {.notes}
Same dataset for the third lecture running, which is deliberate. They fitted the
Michaelis-Menten curve by eye in Lecture 10 with a = 60 and b = 100, and
identified the binomial structure in Lecture 12.

Today those two halves become a generator: a curve for the expected number
killed, and binomial noise around it.
:::

## Simulate the experiment {.smaller}

::: {.type-it}
```{r}
#| fig-height: 2.2
p <- (60 * frogs$Initial / (100 + frogs$Initial)) / frogs$Initial

killed_sim <- rbinom(n = nrow(frogs), size = frogs$Initial, prob = p)

plot(frogs$Initial, killed_sim, pch = 16, col = "red",
     xlab = "tadpoles in the tank", ylab = "number killed")
points(frogs$Initial, frogs$Killed, pch = 1, cex = 1.4)
```
:::

Red: a fake experiment. Open circles: the real one.

::: {.notes}
This is the most important idea in the lecture: if your model is any good, data
simulated from it should be hard to tell from the real thing.

Walk the first line slowly. The curve gives the expected number killed; dividing
by Initial turns that into a per-tadpole probability, which is what rbinom wants.

Run it several times. Sometimes the fake data look like the real data,
sometimes they do not, and the honest conclusion with 16 points is that this
model is not obviously wrong, which is as much as these data support.

In chapter 6 this becomes a formal check. Here it is eyeball.
:::

## What simulation buys you

```{r}
#| echo: false
#| fig-height: 3.0
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
slopes <- replicate(1000, {
  xs <- 1:20
  ys <- rnorm(20, mean = 2 + 1 * xs, sd = 8)
  coef(lm(ys ~ xs))[2]
})
hist(slopes, breaks = 40, col = "grey85", main = "",
     xlab = "estimated slope from 1000 simulated studies")
abline(v = 1, col = "red", lwd = 3)
```

The true slope is **1**. With `sd = 8` and 20 points, individual studies land
anywhere from about 0 to 2.

::: {.notes}
Built with replicate(), which runs an expression many times and collects the
answers. Same idea as sapply from Lecture 9.

This figure is the answer to "why does my result disagree with the published
one". Run the same honest study 1000 times and the estimates spread out this
much. Nobody did anything wrong.

It is also exactly what power analysis counts: how often does a study like this
come back with a slope far enough from zero to notice. Thursday puts a number on
it.

Callback worth making: they saw this figure in Lecture 3 as a finished picture.
Now they can build it.
:::

## Start simple

- Simple models are easier to check and debug
- Complexity adds realism, and new ways to be wrong
- Add one piece at a time, and look after each one

Section **5.2.2.2** has a more complex worked example.

::: {.notes}
The advice is Bolker's and it is the same advice as writing functions in chapter
3: build a small piece, check it, then extend.

Practical version for the midterm: get a line with noise working before adding
groups.
:::

## The midterm {.center}

## The midterm: three parts {.smaller}

**Take-home. Due Tuesday Oct 13, 11:59 pm.** Same dataset you have been using.

| Part | What | Weight |
|---|---|---|
| 1 | A short partial paper: introduction, methods, results with figures | 60% |
| 2 | Given data, decide what model and distribution fit | 30% |
| 3 | A power analysis | 10% |

Any tools you like for code, including AI. **Part 1 text must be yours.**

::: {.notes}
Say the tool policy out loud, because it is unusual and students will assume
otherwise: book, notes, classmates, AI and online resources are all fair for the
coding, because that is what real analysis looks like. The writing in Part 1 is
the exception.

Part 3 needs Thursday's lecture, so nobody should panic about it today.
:::

## What each part uses

**Part 1:** your dataset, Assignment 1 figures, Assignment 2 distributions.
Mostly writing.

**Part 2:** three datasets I give you. Chapter 3 shapes, chapter 4
distributions. This is Assignment 2 with harder data.

**Part 3:** simulate, analyze, repeat, count. How does **noise** change your
chance of detecting a real effect? Today plus Thursday.

. . .

Nothing in it is new. It is the semester so far, assembled.

::: {.notes}
Worth being explicit, because a midterm reads as a wall until you see it is made
of things they have already done.

Part 1 can reuse code and figures from Assignments 1 and 2, and the draft can be
reused again in the final. Say that: the work compounds rather than being thrown
away.

The ESA writing guide is linked in the midterm document for anyone new to
scientific writing.
:::

## Getting started on the midterm

The data and starter code are in the course repository:

```{r}
#| eval: false
library(readr)

twoa <- read_rds("https://github.com/bmaitner/biometry_course/raw/refs/heads/main/midterm/b_2a.RDS")
twob <- read_rds("https://github.com/bmaitner/biometry_course/raw/refs/heads/main/midterm/b_2b.RDS")
twoc <- read_rds("https://github.com/bmaitner/biometry_course/raw/refs/heads/main/midterm/b_2c.RDS")
```

`midterm/midterm_b_example.R` also has the power analysis skeleton for Part 3.

::: {.notes}
Three datasets for Part 2, loaded straight from the repository so nobody has to
download anything. Note the b_ prefix: those are this year's. The unprefixed
2a/2b/2c files are last year's midterm and are still in the folder.

What they are: bird egg mass against adult body mass, reef fish species richness
against total abundance, and hard bottom cover against rugosity.

For each: plot it, decide on a shape and a distribution, justify both. The
procedure is the four steps from Lecture 12.

Also in that script: citation() and citation("packagename"), which Part 1 needs
for citing R and any packages used.
:::

## Remainder of class

Work on the midterm, or keep playing with the simulations.

Good first steps:

- Load the three Part 2 datasets and plot them
- Start the Part 1 introduction, 300 words, on the question you already chose

::: {.notes}
Circulate on Part 2 choices rather than on writing. The shape plus distribution
decision is the part where being in the room helps most.
:::

## Before next class

Read Bolker **5.3**

Next time: **power analysis**, which is Part 3 of the midterm

::: {.notes}
5.3 is short and it is directly needed. Say that the reading is unusually
load-bearing this time: Thursday's class is the method they need for Part 3,
and the midterm is due the Tuesday after.
:::
