---
title: "Probability distributions"
subtitle: "PCB 6455 / BSC 4933: Biometry &nbsp;&nbsp;|&nbsp;&nbsp; Lecture 12"
author: "Dr. Brian Maitner"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    code-line-numbers: false
    fig-width: 7
    fig-height: 4.0
    fig-align: center
    html-math-method: mathml
    include-in-header:
      text: |
        <style>
        .type-it { border-left: 6px solid #2a7ab0; padding-left: 0.4em; }
        .type-it::before, .type-badge {
          content: "Type and run"; display: inline-block;
          background: #2a7ab0; color: white; font-weight: bold;
          font-size: 0.5em; line-height: 1.4; padding: 0 0.6em;
          border-radius: 4px; vertical-align: middle;
        }
        .type-badge { font-size: 0.7em; }
        </style>
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
avonet <- readRDS("../data/Avonet/AVONET1_BirdLife.rds")
bbs <- read.csv("../data/BBS/FL_BBS.csv")
cardinals <- bbs$total_sightings[bbs$Year == 2019 &
                                 bbs$Scientific_Name == "Cardinalis cardinalis"]
library(emdbook)
data(ReedfrogFuncresp)
frogs <- ReedfrogFuncresp
set.seed(1)
```

## Today's agenda

- Assignment 2
- What a probability distribution is
- The four R functions every distribution has
- The bestiary: which distribution for which kind of data
- What the parameters do

. . .

[Type and run]{.type-badge} on a code block: type it into your script and run it.
You will need it later. Code without the label is there to watch.

::: {.notes}
Same convention as the last three lectures.

Today is mostly playing with distributions in R. The reason to do that rather
than read Table 4.1: a distribution is a shape, and you learn a shape by moving
its parameters and watching, not by reading its formula.
:::

## Assignment 2: Distributions

**Already posted. Due Friday Oct 9, by midnight.**

Two parts:

1. **Twelve kinds of data.** For each: discrete or continuous, then name a
   distribution, then one sentence on why.
2. **Your own dataset.** At least three numeric variables: histogram, discrete
   or continuous, mean and median, and an argument for a distribution.

Submit as a `.R` script or the output of a `.Rmd`, same as before.

::: {.notes}
Part 1 answers live in comments if they use a .R file, as text if they use .Rmd.

Say the thing that makes the assignment easier: several items have more than one
defensible answer, and the reasoning is what gets graded. A well argued second
choice beats a bare right answer.

Everything they need for Part 1 is in Table 4.1 and section 4.5, plus today.
:::

## You already have most of this {.smaller}

| The assignment asks for | Where you got it |
|---|---|
| Discrete or continuous | Tuesday |
| Mean, median, and what the gap means | Tuesday |
| `na.rm = TRUE` on real data | Tuesday |
| Histograms, raw and logged | Lecture 5 |
| Which distribution, and why | **today** |
| Checking a distribution against your data | **today** |

Part 2 needs nothing you have not already done.

::: {.notes}
Worth a minute, because an assignment that looks like new work gets started late,
and this one is mostly assembly.

Tuesday did the summary statistics and the discrete versus continuous split.
Today does the distributions themselves and the comparison against data. Lecture
5 did the histograms, including the log scale, which is the step most of them
will need.

The one genuinely new thing is naming a distribution and defending it, which is
what the rest of today is for.
:::

## Where we are

**Chapter 3:** the deterministic part. The curve the data follow on average.

. . .

**Chapter 4:** the stochastic part. How the data scatter around that curve.

. . .

Every model from here on is both: a curve, plus a distribution around it.

::: {.notes}
This is the sentence that makes the chapter make sense. In chapter 3 they fit a
Ricker by eye and the leftover scatter was something to be minimized. Here the
scatter gets its own model.

Bolker's framing: the deterministic function says what you expect, the
distribution says how far from it you land.

Forward reference worth making: in chapter 6 the two get combined, and fitting a
model means choosing both a curve and a distribution.
:::

## What a distribution is

**Discrete:** a probability for each possible value.

$P(X = 3) = 0.18$. The probabilities add to 1.

. . .

**Continuous:** a density, and probability is **area** under it.

$P(X = 3.000...) = 0$. Ask instead for $P(X < 3)$. The total area is 1.

::: {.notes}
From Lecture 11: the exact-value probability for a continuous variable is zero,
so the questions are always about ranges.

The practical consequence shows up in a moment: for discrete distributions
dbinom() gives you an actual probability, and for continuous ones dnorm() gives
you a density, which is not a probability until you integrate it. Nobody needs
to integrate anything today; they need to know why one maxes out above 1.
:::

## Every distribution has four functions {.smaller}

| Letter | Means | Normal | Binomial |
|---|---|---|---|
| `d` | **d**ensity, or probability | `dnorm()` | `dbinom()` |
| `p` | **p**robability of being below | `pnorm()` | `pbinom()` |
| `q` | **q**uantile, the value at a probability | `qnorm()` | `qbinom()` |
| `r` | **r**andom draws | `rnorm()` | `rbinom()` |

The pattern holds for every distribution in section 4.5.

::: {.notes}
This table is the single most useful thing on the slides today. Once they have
it, every distribution in the bestiary is four functions they already know how
to call.

Point at the pattern rather than the entries: the letter says what you want, the
name says which distribution. dpois, ppois, qpois, rpois. dgamma, pgamma, and so
on.

?dnorm shows all four on one help page, which is worth demonstrating live.
:::

## `d` is for density

::: {.type-it}
```{r}
#| fig-height: 3.0
x_vector <- seq(from = -10, to = 10, by = 0.1)

plot(dnorm(x = x_vector, mean = 0, sd = 1) ~ x_vector)
```
:::

::: {.notes}
Deliberately their old slide's code, because it works and because the formula
interface `y ~ x` was just met in Lecture 10.

Note the shape of the call: a vector of x values goes in, a vector of heights
comes out. Same idea as the functions they wrote in chapter 3, which is the
point: a density is just a function.

Anyone who wants a line instead of circles: add type = "l".
:::

## Your turn: what do the parameters do? {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
x_vector <- seq(from = -10, to = 10, by = 0.1)

# Round 1: change the mean, keep sd the same
plot(dnorm(x = x_vector, mean = 0, sd = 1) ~ x_vector, type = "l")
plot(dnorm(x = x_vector, mean = 3, sd = 1) ~ x_vector, type = "l")

# Round 2: keep the mean, change sd
plot(dnorm(x = x_vector, mean = 0, sd = 2) ~ x_vector, type = "l")
plot(dnorm(x = x_vector, mean = 0, sd = 4) ~ x_vector, type = "l")
```
:::

In your own words: what does `mean` do? What does `sd` do?

::: {.notes}
Five minutes. Same move as the negative exponential in Lecture 9: change only
the numbers.

Watch for the thing that makes this worth doing live: each plot gets its own y
axis, so the sd = 4 curve looks the same height as the sd = 1 curve until you
read the axis. Fixing it with ylim = c(0, 0.4) is the lesson from Lecture 6
about comparing panels, met again.

Answers to listen for: the mean slides the curve left and right without changing
its shape, the sd stretches it. A scale parameter and a location parameter,
which is the vocabulary from chapter 3.
:::

## What they do

```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 9
x_vector <- seq(-10, 10, by = 0.05)
par(mfrow = c(1, 2), mar = c(3.2, 3.4, 2.4, 0.8), mgp = c(2, 0.6, 0))
plot(dnorm(x_vector, mean = 0, sd = 1) ~ x_vector, type = "l", lwd = 3,
     col = "red", ylim = c(0, 0.45), xlab = "x", ylab = "density",
     main = "Changing the mean")
lines(x_vector, dnorm(x_vector, mean = 3, sd = 1), lwd = 3, col = "blue")
lines(x_vector, dnorm(x_vector, mean = -4, sd = 1), lwd = 3, col = "grey50")
legend("topleft", bty = "n", lwd = 3, col = c("grey50", "red", "blue"),
       legend = c("-4", "0", "3"), cex = 0.9)
plot(dnorm(x_vector, mean = 0, sd = 1) ~ x_vector, type = "l", lwd = 3,
     col = "red", ylim = c(0, 0.45), xlab = "x", ylab = "density",
     main = "Changing the sd")
lines(x_vector, dnorm(x_vector, mean = 0, sd = 2), lwd = 3, col = "blue")
lines(x_vector, dnorm(x_vector, mean = 0, sd = 4), lwd = 3, col = "grey50")
legend("topleft", bty = "n", lwd = 3, col = c("red", "blue", "grey50"),
       legend = c("1", "2", "4"), cex = 0.9)
```

`mean` moves it. `sd` spreads it. The area stays 1.

::: {.notes}
On the same axes, which is what their four separate plots could not show.

The area point explains why the wide curve is shorter: total probability is
fixed at 1, so spreading it out lowers it. That is the difference between a
density and a probability showing up visually.
:::

## `p` is for probability below

```{r}
#| fig-height: 2.6
pnorm(q = 0, mean = 0, sd = 1)
pnorm(q = 1.96, mean = 0, sd = 1)
```

```{r}
#| echo: false
#| fig-height: 2.4
par(mar = c(3.2, 3.2, 0.5, 0.8), mgp = c(1.9, 0.6, 0))
xs <- seq(-4, 4, by = 0.01)
plot(dnorm(xs) ~ xs, type = "l", lwd = 2, xlab = "x", ylab = "density")
shade <- xs[xs <= 1.96]
polygon(c(shade, 1.96), c(dnorm(shade), 0), col = "lightsteelblue", border = NA)
lines(xs, dnorm(xs), lwd = 2)
```

The shaded area is 0.975.

::: {.notes}
Half the area is below the mean, which is the first line and a good sanity check.

1.96 is worth naming now, quietly: 97.5% of a normal distribution is below it,
so 95% sits between -1.96 and 1.96. They will meet that number constantly from
chapter 6 on, and meeting it first as an area is better than meeting it as a
magic constant.

This is also the answer to "what is a p-value" at the level they need today:
an area under a distribution.
:::

## `q` is for quantile

`q` is `p` backwards. Give it a probability, it gives you the value.

```{r}
qnorm(p = 0.975, mean = 0, sd = 1)
qnorm(p = 0.5, mean = 10, sd = 2)
```

::: {.notes}
Point at the pair: pnorm(1.96) is 0.975 and qnorm(0.975) is 1.96. Same
relationship read in the two directions.

The second line is a free check on understanding: the median of a normal
distribution is its mean.

Where they will use it: confidence intervals in chapter 6 are built from
quantiles.
:::

## `r` is for random

::: {.type-it}
```{r}
set.seed(42)
rnorm(n = 5, mean = 0, sd = 1)
```
:::

. . .

`set.seed()` makes the randomness repeatable. Run it again, same five numbers.

::: {.notes}
This is the function they will use most for the rest of the semester, and it is
the whole of chapter 5.

set.seed() is worth the thirty seconds: without it a script that simulates gives
different answers each run, which makes it hard to tell a real change from
noise. With it, the work is reproducible, which is the Assignment 2 criterion
and the Lecture 27 topic.

Ask what happens without the seed. Run it twice.
:::

## Your turn: simulate and compare {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
draws <- rnorm(n = 100, mean = 10, sd = 2)

hist(draws, freq = FALSE)

curve(dnorm(x, mean = 10, sd = 2), add = TRUE, col = "red", lwd = 2)
```
:::

Now change `n` to 10000. What happens to the match?

::: {.notes}
Six minutes. This slide is the bridge between a distribution as an idea and data
as something you hold.

freq = FALSE is the key argument: it puts the histogram on the density scale so
the curve and the bars are comparable. With freq = TRUE the curve is invisible
along the bottom, which someone will see.

curve() is back from chapter 3, with add = TRUE from Lecture 9.

The lesson from raising n: the histogram wobbles around the curve, and the wobble
shrinks as n grows. The distribution is what the data look like in the limit.
That is also why 20 data points cannot tell you much about a distribution's
shape, which is the honest caveat for Assignment 2 Part 2.
:::

## The bestiary {.center}

## How do you choose?

1. **Discrete or continuous?** Counts or measurements.
2. **What range?** 0 to 1, 0 to infinity, any number.
3. **What process generated it?** Table 4.1 is organized this way.
4. **Compare the shape** to a histogram of your data.
5. **Look at what others did** with data like yours.

. . .

Then **justify it**. More than one answer can be defensible.

::: {.notes}
Their old slide, kept, because it is the procedure Assignment 2 Part 1 is
testing. Steps 1 and 2 eliminate most of Table 4.1 in about ten seconds.

Step 3 is the one that separates a good answer from a guess, and it is what the
next few slides are about: each distribution comes from a story about how the
numbers were made.
:::

## Discrete: binomial

:::: {.columns}

::: {.column width="50%"}
**The story:** a fixed number of tries, each succeeds or fails independently
with the same probability.

- `size`: number of tries
- `prob`: probability of success

**Range:** 0 to `size`

`dbinom()`, `rbinom()`
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.6
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
k <- 0:20
plot(k, dbinom(k, size = 20, prob = 0.5), type = "h", lwd = 4, col = "red",
     xlab = "number of successes", ylab = "probability", ylim = c(0, 0.3))
points(k, dbinom(k, size = 20, prob = 0.2), type = "h", lwd = 4,
       col = adjustcolor("blue", 0.6))
legend("topright", bty = "n", lwd = 4, col = c("blue", "red"),
       legend = c("prob = 0.2", "prob = 0.5"), cex = 0.85)
```
:::

::::

::: {.notes}
Examples from the assignment: survivors out of a known starting number, larvae
settling out of larvae released. The giveaway phrase is "out of".

The ceiling matters. A binomial can never exceed size, which is what separates
it from the count distributions on the next slide.
:::

## Your turn: the binomial {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
k <- seq(from = 0, to = 20, by = 1)

plot(dbinom(x = k, size = 20, prob = 0.5) ~ k, ylim = c(0, 0.3))

# now try these, one at a time
# size = 20, prob = 0.1
# size = 20, prob = 0.9
# size = 5,  prob = 0.5
```
:::

What does `prob` do? What does `size` do?

::: {.notes}
Five minutes.

Answers: prob slides the peak left or right, since the expected number of
successes is size times prob. size sets how far the distribution can reach and
also widens it.

Worth pointing out while circulating: with prob = 0.9 the distribution is skewed
toward the top, which is the first asymmetric distribution they have drawn.

The ylim is there on purpose, from the Lecture 6 lesson about comparable axes.
:::

## Discrete: Poisson

:::: {.columns}

::: {.column width="50%"}
**The story:** counts of things that happen independently at a constant average
rate, with no ceiling.

- `lambda`: the mean count

**Range:** 0 to infinity

Mean and variance are **both** `lambda`.

`dpois()`, `rpois()`
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-height: 3.6
#| fig-width: 4.6
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
k <- 0:20
plot(k, dpois(k, lambda = 1), type = "h", lwd = 4, col = "grey40",
     xlab = "count", ylab = "probability", ylim = c(0, 0.4))
points(k + 0.25, dpois(k, lambda = 3), type = "h", lwd = 4, col = "red")
points(k + 0.5, dpois(k, lambda = 8), type = "h", lwd = 4, col = "blue")
legend("topright", bty = "n", lwd = 4, col = c("grey40", "red", "blue"),
       legend = c("lambda = 1", "lambda = 3", "lambda = 8"), cex = 0.85)
```
:::

::::

::: {.notes}
Seeds per quadrat, visits per flower per hour, fish on a transect. The giveaway
is a count with no natural maximum.

The mean equals the variance claim is the testable part, and it is what fails on
real data constantly, which sets up the next slide.

Notice the shape changes as lambda grows: skewed at lambda = 1, nearly
symmetric by lambda = 8. A Poisson with a large mean looks normal, which is
worth saying because it explains why normal approximations show up everywhere.
:::

## When the Poisson is not enough

```{r}
#| echo: false
#| fig-height: 3.4
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
k <- 0:25
plot(k, dpois(k, lambda = 5), type = "h", lwd = 5, col = "grey50",
     xlab = "count", ylab = "probability", ylim = c(0, 0.25))
points(k + 0.3, dnbinom(k, mu = 5, size = 1), type = "h", lwd = 5, col = "red")
legend("topright", bty = "n", lwd = 5, col = c("grey50", "red"),
       legend = c("Poisson, mean 5", "negative binomial, mean 5"))
```

Same mean. The **negative binomial** allows more zeros and more big values:
**clumping**.

::: {.notes}
Real organisms are not spread at random. They cluster, around resources, around
each other, or because conditions vary from plot to plot. Clumping makes the
variance exceed the mean, which a Poisson cannot do.

This is overdispersion, and it is the single most common reason a count model
fails. dnbinom() takes mu and size, where size controls the clumping: small size
means heavy clumping, and as size grows it becomes a Poisson.

For Assignment 2, "Poisson, unless individuals cluster, in which case negative
binomial" is the best answer a student can give on a count item.
:::

## Two more discrete

:::: {.columns}

::: {.column width="50%"}
**Geometric**

How many tries until the first success.

- `prob`: success probability

Ages, if the chance of dying each year is constant.

`dgeom()`
:::

::: {.column width="50%"}
**Beta-binomial**

A binomial whose `prob` itself varies between groups.

Infection prevalence across ponds, settlement across tanks.

`dbetabinom()` in `emdbook`
:::

::::

```{r}
#| echo: false
#| fig-height: 2.4
#| fig-width: 9
par(mfrow = c(1, 2), mar = c(3.2, 3.4, 1.6, 0.6), mgp = c(1.9, 0.6, 0))
k <- 0:20
plot(k, dgeom(k, prob = 0.3), type = "h", lwd = 4, col = "red",
     xlab = "tries before success", ylab = "probability", main = "Geometric")
plot(k, emdbook::dbetabinom(k, prob = 0.5, size = 20, theta = 5), type = "h",
     lwd = 4, col = "red", xlab = "successes", ylab = "probability",
     ylim = c(0, 0.2), main = "Beta-binomial vs binomial")
points(k + 0.3, dbinom(k, size = 20, prob = 0.5), type = "h", lwd = 4,
       col = "grey60")
```

::: {.notes}
Both are in the bestiary and both show up on Assignment 2.

Geometric is the discrete version of the exponential, which is on the next
slide. Constant hazard, so most waits are short and a few are long.

Beta-binomial is to the binomial what the negative binomial is to the Poisson:
the same story plus variation in the rate. Say that pairing out loud, it makes
four distributions into two ideas.
:::

## Continuous: normal and lognormal

:::: {.columns}

::: {.column width="50%"}
**Normal**

Many small effects added together.

Range: any number, including negatives.

`dnorm()`, `rnorm()`
:::

::: {.column width="50%"}
**Lognormal**

Many small effects **multiplied** together.

Range: positive only. Right skewed.

Normal after a log transform.

`dlnorm()`, `rlnorm()`
:::

::::

```{r}
#| echo: false
#| fig-height: 2.6
#| fig-width: 9
par(mfrow = c(1, 2), mar = c(3.2, 3.4, 1.6, 0.6), mgp = c(1.9, 0.6, 0))
xs <- seq(-4, 4, by = 0.01)
plot(xs, dnorm(xs), type = "l", lwd = 3, col = "red", xlab = "x",
     ylab = "density", main = "Normal")
xs2 <- seq(0.001, 8, by = 0.01)
plot(xs2, dlnorm(xs2, meanlog = 0, sdlog = 1), type = "l", lwd = 3, col = "red",
     xlab = "x", ylab = "density", main = "Lognormal")
```

::: {.notes}
The added versus multiplied distinction is the most useful sentence in the
chapter. Height is many small additive contributions. Body mass, population
size, and most abundances are multiplicative, each year's value a factor times
the last, which produces the long right tail.

That is why log transforming works so often in biology, and why Lecture 5 used
log axes for mass. It is not a trick to make a plot look nicer; it is a claim
about how the numbers were made.

Negative values rule out the lognormal instantly, which is a fast
elimination step for Assignment 2.
:::

## Continuous: waiting times and proportions

:::: {.columns}

::: {.column width="50%"}
**Exponential and gamma**

Time or distance until something happens.

Exponential: constant hazard.
Gamma: the wait for several events.

`dexp()`, `dgamma()`
:::

::: {.column width="50%"}
**Beta**

A proportion, between 0 and 1.

Cover, prevalence, any fraction.

`dbeta()`
:::

::::

```{r}
#| echo: false
#| fig-height: 2.6
#| fig-width: 9
par(mfrow = c(1, 2), mar = c(3.2, 3.4, 1.6, 0.6), mgp = c(1.9, 0.6, 0))
xs <- seq(0, 8, by = 0.01)
plot(xs, dexp(xs, rate = 1), type = "l", lwd = 3, col = "red", ylim = c(0, 1),
     xlab = "x", ylab = "density", main = "Exponential and gamma")
lines(xs, dgamma(xs, shape = 2, rate = 1), lwd = 3, col = "blue")
lines(xs, dgamma(xs, shape = 5, rate = 1), lwd = 3, col = "grey40")
legend("topright", bty = "n", lwd = 3, col = c("red", "blue", "grey40"),
       legend = c("exp", "gamma, shape 2", "gamma, shape 5"), cex = 0.8)
ps <- seq(0.001, 0.999, by = 0.001)
plot(ps, dbeta(ps, 2, 5), type = "l", lwd = 3, col = "red", ylim = c(0, 3),
     xlab = "proportion", ylab = "density", main = "Beta")
lines(ps, dbeta(ps, 5, 2), lwd = 3, col = "blue")
lines(ps, dbeta(ps, 1, 1), lwd = 3, col = "grey40")
legend("top", bty = "n", lwd = 3, col = c("red", "blue", "grey40"),
       legend = c("beta(2,5)", "beta(5,2)", "beta(1,1) = uniform"), cex = 0.8)
```

::: {.notes}
The exponential is the gamma with shape 1, which the figure shows. As shape
grows the gamma stops being highest at zero and develops a hump, because waiting
for five events is unlikely to be quick.

Time to germination, time to death, step lengths: all waiting times.

Beta covers the 0 to 1 range, and beta(1,1) is flat, the uniform. Worth pointing
at that line: the uniform is not a separate idea, it is a beta with both
parameters 1.
:::

## The whole bestiary, in one picture

```{r}
#| echo: false
#| fig-width: 9.5
#| fig-height: 5
par(mfrow = c(2, 3), mar = c(2.6, 2.6, 2.2, 0.6), mgp = c(1.4, 0.5, 0))
k <- 0:20
plot(k, dbinom(k, 20, 0.3), type = "h", lwd = 4, col = "red",
     xlab = "", ylab = "", main = "Binomial\nout of a fixed number")
plot(k, dpois(k, 4), type = "h", lwd = 4, col = "red",
     xlab = "", ylab = "", main = "Poisson\ncounts, no ceiling")
plot(k, dnbinom(k, mu = 4, size = 1), type = "h", lwd = 4, col = "red",
     xlab = "", ylab = "", main = "Negative binomial\nclumped counts")
xs <- seq(-4, 4, by = 0.01)
plot(xs, dnorm(xs), type = "l", lwd = 3, col = "red",
     xlab = "", ylab = "", main = "Normal\nadditive")
xs2 <- seq(0.001, 8, by = 0.01)
plot(xs2, dlnorm(xs2), type = "l", lwd = 3, col = "red",
     xlab = "", ylab = "", main = "Lognormal\nmultiplicative")
ps <- seq(0.001, 0.999, by = 0.001)
plot(ps, dbeta(ps, 2, 5), type = "l", lwd = 3, col = "red",
     xlab = "", ylab = "", main = "Beta\nproportions")
```

::: {.notes}
The six that cover most of what they will meet, and most of Assignment 2.

Top row discrete, bottom row continuous. Worth leaving up while they do the next
exercise.

Table 4.1 has the rest, including the geometric and the gamma.
:::

## Your turn: name a distribution {background-color="#f0f4f8" .smaller}

No code. Discrete or continuous first, then a distribution, then why.

1. Number of seeds germinating out of 50 planted
2. Number of beetles on each of 200 trees, where beetles aggregate
3. Time until a radio-tagged bird leaves the study area
4. Proportion of each quadrat covered by moss
5. Body mass of adult fish
6. Number of lightning strikes per square kilometer per year

::: {.notes}
Eight minutes in pairs. This is Assignment 2 Part 1 with training wheels, and it
is the best predictor of how that part will go.

Answers: 1 binomial, out of a fixed number. 2 negative binomial, the word
aggregate is the tell. 3 exponential or gamma, a waiting time. 4 beta, a
proportion. 5 lognormal, positive and right skewed. 6 Poisson, counts per unit
area and time with no ceiling.

Push on the why rather than the name. A student who says "Poisson because counts
with no ceiling" has the method; a student who says "Poisson because it is a
count" will pick Poisson for item 2 as well.
:::

## Does it match the data?

```{r}
#| fig-height: 3.2
hist(log10(avonet$Mass), freq = FALSE, breaks = 40,
     main = "", xlab = "log10 body mass (g)")

curve(dnorm(x, mean = mean(log10(avonet$Mass)), sd = sd(log10(avonet$Mass))),
      add = TRUE, col = "red", lwd = 3)
```

Roughly **lognormal**: much closer to normal once you take the log, but still
right skewed. **Close enough to be useful is the standard.**

::: {.notes}
The payoff slide, and the model for Assignment 2 Part 2. Three steps: histogram
on the density scale, overlay the candidate distribution, look.

Be honest about the fit, and make that the lesson. The curve sits to the right
of the data: there are more small birds than a normal distribution on the log
scale predicts, and the skew on the log scale is still about 0.85. This is
11,000 species from many families pooled together, so no single simple
distribution should fit perfectly.

The useful question is not "is it exactly lognormal", which is always no, but
"is lognormal a better description than normal", which is clearly yes: plot the
raw mass and the curve is hopeless.

Worth asking the room what would make this fit better. Splitting by family or by
trophic level is the honest answer, and that is what the rest of the course is
for.

freq = FALSE again, for the same reason as the simulation slide.

This is also the answer to "how did you know it was lognormal?" from Lecture 5,
when we logged the mass axis without justifying it.
:::

## Real data, three times {.center}

## Example 1: counting cardinals

Breeding Bird Survey, Florida, 2019. One count per route, 71 routes.

```{r}
#| echo: false
#| fig-height: 3.0
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
hist(cardinals, breaks = seq(0, 180, by = 10), col = "grey85",
     main = "", xlab = "Northern cardinals counted on a route")
```

A count with no ceiling. **Poisson?**

::: {.notes}
Real data in the course repository, from `data/BBS/`, and a dataset several of
them may use for Assignment 2.

One row per route: a volunteer stops every half mile along a 24.5 mile route and
counts every bird heard or seen. So the count is "cardinals detected on this
route this year".

Run the shape test from earlier: counts, no fixed ceiling, so the bestiary says
Poisson. The next slide checks that properly rather than taking it on faith.
:::

## The Poisson makes a testable claim

::: {.type-it}
```{r}
mean(cardinals)
var(cardinals)
var(cardinals) / mean(cardinals)
```
:::

A Poisson has **mean = variance**, so that ratio should be about 1.

It is **28**.

::: {.notes}
This is the most useful two lines of code in the lecture, and it takes ten
seconds on any count data they will ever have.

The ratio is the diagnostic. Around 1 means a Poisson is plausible. Much greater
than 1 means overdispersion, which is almost always what real ecological counts
look like.

Why cardinals clump: habitat varies between routes, some routes are suburban and
some are pine flatwoods, and birds respond to that. Variation in the rate between
routes is exactly the negative binomial story.
:::

## How wrong is the Poisson here?

```{r}
#| echo: false
#| fig-height: 3.2
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
mu <- mean(cardinals); v <- var(cardinals); sz <- mu^2 / (v - mu)
k <- 0:180
hist(cardinals, breaks = seq(0, 180, by = 10), freq = FALSE, col = "grey88",
     border = "grey60", main = "", xlab = "cardinals per route",
     ylim = c(0, 0.062))
lines(k, dpois(k, lambda = mu), col = "blue", lwd = 3)
lines(k, dnbinom(k, mu = mu, size = sz), col = "red", lwd = 3)
legend("topright", bty = "n", lwd = 3, col = c("blue", "red"),
       legend = c("Poisson", "negative binomial"))
```

One route counted **175**. The Poisson says that is essentially impossible
($10^{-43}$). The negative binomial says about 1 in 100.

::: {.notes}
The blue curve is too narrow and too tall: a Poisson with mean 49 expects almost
everything between 35 and 65, and the data run from single digits to 175.

The 10^-43 number is the honest way to say "wrong model". It is not that the
Poisson fits slightly worse, it is that it calls the observed data impossible.

Where the red curve comes from is the next slide, and it is one line of
arithmetic.
:::

## Matching a negative binomial

The negative binomial has a mean and a clumping parameter. Solve for them from
the data:

::: {.type-it}
```{r}
mu <- mean(cardinals)
size <- mu^2 / (var(cardinals) - mu)

mu
size
```
:::

Small `size` means heavy clumping. As `size` grows it becomes a Poisson.

::: {.notes}
This is the method of moments: set the distribution's mean and variance equal to
the data's, then solve. It is the simplest way to fit a distribution and it
needs no optimizer.

Chapter 6 replaces it with maximum likelihood, which is better but no more
honest. Worth saying so now: they have just fitted a distribution to real data,
in two lines, in week five.

size = 1.8 is heavy clumping. For reference, size above about 10 would look
nearly Poisson at this mean.
:::

## Your turn: a different species {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
bbs <- read.csv("https://github.com/bmaitner/biometry_course/raw/refs/heads/main/data/BBS/FL_BBS.csv")

counts <- bbs$total_sightings[bbs$Year == 2019 &
                              bbs$Scientific_Name == "Melanerpes carolinus"]

hist(counts)

mean(counts)
var(counts) / mean(counts)
```
:::

Is this one clumped too? Try another species of your own.

::: {.notes}
Six minutes. The red-bellied woodpecker gives a ratio of about 5.3: still
overdispersed, but much less than the cardinal's 28.

That contrast is the point. Overdispersion is not all or nothing, and the ratio
puts a number on how badly a Poisson would do.

Species names in the data are scientific names. Mourning dove, Zenaida macroura,
gives about 16. Anyone who picks a rare species will get few routes and a noisy
answer, which is itself worth discussing.
:::

## Example 2: killed out of how many

Tadpole predation. Each tank started with a known number, so each result is
"killed **out of** Initial".

```{r}
#| echo: false
#| fig-height: 2.8
par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2, 0.6, 0))
plot(frogs$Initial, frogs$Killed / frogs$Initial, pch = 16, cex = 1.5,
     ylim = c(0, 0.8), xlab = "tadpoles in the tank",
     ylab = "proportion killed")
```

Out of a fixed number, so **binomial**. But the proportion is not constant.

::: {.notes}
Same data as Lecture 10, asked a different way. There the question was how many
were killed; here it is what fraction.

The giveaway phrase for a binomial is "out of", and it is on the Assignment 2
list twice.

Read the plot: the proportion killed is around 0.5 at low density and drops to
about 0.3 at high density. A single binomial with one p cannot produce that,
because p would have to change with density.
:::

## This is the whole course in one slide

**Chapter 3:** a curve for how the proportion changes with density.

**Chapter 4:** a binomial for the scatter around that curve.

. . .

**Chapter 6:** fit both at once. That is what a model is.

::: {.notes}
Worth slowing down for. They have now built both halves separately, five weeks
apart, on the same dataset.

The curve is the functional response from Lecture 10, the Holling type II, which
they fitted by eye. The noise is today's binomial. Putting them together is
chapter 6, and when it arrives it should feel like assembly rather than
novelty.

If anyone asks what "fit both at once" means: choose the curve's parameters so
that the binomial scatter around it makes the observed data as likely as
possible. That sentence is maximum likelihood.
:::

## Your turn: how big are bird ranges? {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
r <- avonet$Range.Size

hist(r, breaks = 50)
mean(r, na.rm = TRUE)
median(r, na.rm = TRUE)

hist(log10(r), breaks = 50)
```
:::

Mean 2,580,000 km². Median 416,000 km². What does that gap tell you?
Does the log fix it?

::: {.notes}
Six minutes, and the answer is more interesting than it looks.

The raw variable is hugely right skewed, which the mean against the median says
before any plot: the mean is six times the median.

The log10 version is better but overshoots, leaving a left skew of about -0.95.
So this one is not lognormal either. The right answer to "which distribution" is
"none of these cleanly", and saying that with evidence is a full credit answer on
Assignment 2.

Also note the 57 missing values, which is why na.rm is in the code.
:::

## What the three examples had in common

1. **Plot it.** Shape first.
2. **Use a number, not an impression.** Mean against variance for counts, mean
   against median for skew.
3. **Overlay the candidate** on the data.
4. **Say how it fails**, and whether that matters.

Nobody looked up a formula.

::: {.notes}
This is the procedure for Assignment 2 Part 2, and it is four steps they have now
done three times on real data.

Step 4 is the one that separates a good answer from a confident wrong one. Every
distribution is wrong about something; the question is whether it is wrong about
the thing you care about.
:::

## A warning: groups hide inside your data

```{r}
#| echo: false
#| fig-height: 3.0
#| fig-width: 9
set.seed(7)
males   <- rlnorm(300, meanlog = log(60), sdlog = 0.18)
females <- rlnorm(300, meanlog = log(95), sdlog = 0.18)
par(mfrow = c(1, 2), mar = c(3.4, 3.4, 2, 0.6), mgp = c(2, 0.6, 0))
hist(c(males, females), breaks = 30, col = "grey85",
     main = "All lizards measured", xlab = "body length (mm)")
hist(males, breaks = 25, col = rgb(0.2, 0.4, 1, 0.55), ylim = c(0, 30),
     xlim = range(c(males, females)), main = "Split by sex",
     xlab = "body length (mm)")
hist(females, breaks = 25, col = rgb(1, 0.3, 0.1, 0.5), add = TRUE)
legend("topright", fill = c(rgb(0.2, 0.4, 1, 0.55), rgb(1, 0.3, 0.1, 0.5)),
       legend = c("males", "females"), bty = "n")
```

Each sex is **lognormal**. Pooled, they look like nothing in the bestiary.

::: {.notes}
Simulated, so the right answer is known: both groups are lognormal, the females
larger. Pooled, the histogram is wide and lumpy and matches no standard shape.

This is the most common reason a real variable refuses to look like anything in
Table 4.1, and it is not a flaw in the data. It means the sample contains more
than one kind of thing.

Sexual size dimorphism is the obvious case in lizards. Others they will hit:
species pooled together, sites with different conditions, juveniles and adults,
sampling years.
:::

## The same thing in real data

```{r}
#| echo: false
#| fig-height: 3.0
#| fig-width: 9
amniote <- readRDS("../data/Amniote_traits/Amniote_Database_Aug_2015.rds")
amniote[amniote == -999] <- NA
m <- amniote[!is.na(amniote$adult_body_mass_g) &
             amniote$class %in% c("Aves", "Mammalia"), ]
x <- log10(m$adult_body_mass_g)
par(mfrow = c(1, 2), mar = c(3.4, 3.4, 2, 0.6), mgp = c(2, 0.6, 0))
hist(x, breaks = 45, col = "grey85", main = "Birds and mammals pooled",
     xlab = "log10 body mass (g)")
hist(x[m$class == "Aves"], breaks = 40, col = rgb(0.2, 0.4, 1, 0.55),
     xlim = range(x), main = "Split by class", xlab = "log10 body mass (g)")
hist(x[m$class == "Mammalia"], breaks = 40, col = rgb(1, 0.3, 0.1, 0.5),
     add = TRUE)
legend("topright", fill = c(rgb(0.2, 0.4, 1, 0.55), rgb(1, 0.3, 0.1, 0.5)),
       legend = c("birds", "mammals"), bty = "n")
```

The long right tail is mammals. Birds alone are far tighter.

::: {.notes}
Amniote trait database, in the course repository.

Numbers if useful: on the log10 scale the pooled skew is 1.5, birds alone 0.8.
Birds have a standard deviation of 0.69 and mammals 1.26, so mammals are nearly
twice as spread out. The second bump around 10,000 to 100,000 g is large
mammals, which have no bird equivalent.

Honest point: splitting does not make these perfectly lognormal either. It makes
them better, and it tells you something biological, which is the real payoff.
:::

## What to do about it {.smaller}

1. **Plot by group before naming a distribution.** `boxplot(y ~ group)`, or one
   histogram per group.
2. **Suspect a mixture** when you see two humps, a shoulder, or a spread far
   wider than any single distribution would give.
3. **If the groups differ, describe them separately.** A distribution per group
   beats one bad distribution for everything.
4. **If you cannot measure the group, say so.** That is an honest answer, and it
   is section 4.6.

. . .

Later in the course the group stops being a problem and becomes a **predictor**.

::: {.notes}
Point 3 is what Assignment 2 Part 2 wants when a variable is structured: not a
shrug, and not a forced label, but "males and females each look lognormal with
different means".

Point 4 covers the genuinely hard cases. Unmeasured structure is why zero
inflation and mixture models exist, and 4.6 is where Bolker puts them.

The last line is the forward reference that makes this feel less like a warning
and more like a preview: from chapter 9 on, sex or site goes into the model as a
term, and the "mixture" becomes the thing being estimated.
:::

## Your turn: is your variable one thing? {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
# a grouping column: sex, site, species, year, treatment
boxplot(my_data$your_variable ~ my_data$your_group)

# or one histogram per group
hist(my_data$your_variable[my_data$your_group == "A"])
hist(my_data$your_variable[my_data$your_group == "B"])
```
:::

Does the picture change? If it does, say so in Assignment 2.

::: {.notes}
Four minutes, and it rolls straight into the Part 2 work block.

Most of their datasets have a grouping column they have not thought about:
species, site, sex, year, treatment. Asking "is this one thing?" once is usually
enough to find it.

If a student has no grouping variable at all, that is worth saying out loud too.
A single species measured once in one place is the easy case.
:::

## Your turn: start Assignment 2, Part 2 {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
my_data <- read.csv("your_file.csv")

x <- my_data$your_variable

hist(x, freq = FALSE, breaks = 30)

mean(x, na.rm = TRUE)
median(x, na.rm = TRUE)

# if it is positive and right skewed, try the log
hist(log10(x), freq = FALSE, breaks = 30)
```
:::

Then: discrete or continuous? Which distribution, and why?

::: {.notes}
The rest of class. This is Part 2 of the assignment, started with help in the
room, which is the point of doing it here.

Circulate on the decision, not the code. The question to ask each student is
"what process made this number?", which is step 3 of the how-do-you-choose
slide.

Common cases to expect: counts stored as numeric, variables with a hard upper
bound that make a proportion, and heavily zero-inflated data, which is genuinely
hard and worth saying so. Zero inflation is in section 4.6.
:::

## Remainder of class

Work on **Assignment 2**.

- Part 1: the twelve items, using Table 4.1 and section 4.5
- Part 2: your own data, the three steps above

Due **Friday Oct 9**, by midnight.

::: {.notes}
Point out that Part 1 is doable now: today covered every distribution on the
list.
:::

## Before next class

Read Bolker **5.1 to 5.2**

Next time: stochastic simulation, and the midterm explained

::: {.notes}
Chapter 5 is where rnorm and friends stop being a demonstration and become a
method: simulating data to find out what your analysis would do before you
collect anything.
:::
