---
title: "Reef exercise: worked solution"
subtitle: "PCB 6455 / BSC 4933: Biometry &nbsp;&nbsp;|&nbsp;&nbsp; Lecture 14"
author: "Dr. Brian Maitner"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    code-line-numbers: false
    fig-width: 7
    fig-height: 4.0
    fig-align: center
    html-math-method: mathml
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
knitr::knit_hooks$set(small_mar = function(before, options, envir) {
  if (before) par(mar = c(3.4, 3.4, 0.6, 0.6), mgp = c(2.1, 0.6, 0))
})
knitr::opts_chunk$set(small_mar = TRUE)
set.seed(3)
```

## How to use this

One possible answer to the reef exercise, with the code filled in.

. . .

**Your answers will differ**, and that is fine. Two places where a different
choice is just as defensible are flagged as we go.

What matters is that each decision was made for a reason you could state.

::: {.notes}
Posted after class so students can check their own work rather than wonder.
:::

## Setting up

```{r}
reef <- read.csv("https://github.com/bmaitner/biometry_course/raw/refs/heads/main/data/Reef_fish/NCRMP_reef_fish_surveys.csv")

reef$protected <- ifelse(reef$PROT == 0, "no", "yes")

# 8 of 8560 surveys recorded no fish, and log(0) is -Inf
reef <- reef[reef$total_abundance > 0, ]

unprot <- reef$total_abundance[reef$protected == "no"]
prot   <- reef$total_abundance[reef$protected == "yes"]

c(unprotected = length(unprot), protected = length(prot))
```

## 1. The deterministic part {.smaller}

```{r}
#| fig-height: 2.3
#| fig-width: 9
par(mfrow = c(1, 2))
hist(unprot, breaks = 40, main = "raw", xlab = "fish per survey")
hist(log(unprot), breaks = 40, main = "logged", xlab = "log fish per survey")
```

Raw is strongly right skewed, logged is near symmetric. So describe the effect
as a **ratio**, not a difference.

```{r}
c(median_unprot = median(unprot), median_prot = median(prot),
  ratio = median(prot) / median(unprot))
```

::: {.notes}
A ratio also travels better: "50% more fish" means the same thing at any reef,
while "73 more fish per survey" does not.
:::

## 2. The distribution {.smaller}

Positive, continuous, right skewed, symmetric once logged: **lognormal**.
Gamma would also have been defensible.

The check, on something we did not fit:

```{r}
mean(unprot > 500)                                     # observed

plnorm(500, meanlog = mean(log(unprot)),
       sdlog = sd(log(unprot)), lower.tail = FALSE)    # predicted
```

Same ballpark. The lognormal slightly **over-predicts** very rich surveys, which
is worth noticing rather than hiding.

::: {.notes}
Anyone who chose gamma and checked it the same way has done the exercise
correctly. The checking is the skill, not the answer.
:::

## 3. The parameters

```{r}
meanlog_unprot <- mean(log(unprot))
sdlog_unprot   <- sd(log(unprot))

meanlog_prot <- mean(log(prot))
sdlog_prot   <- sd(log(prot))

round(c(meanlog_unprot = meanlog_unprot, sdlog_unprot = sdlog_unprot,
        meanlog_prot = meanlog_prot, sdlog_prot = sdlog_prot), 2)
```

The two spreads are nearly identical, so one spread can be used for both groups.

`meanlog` is about 5 because `exp(5)` is about 150 fish. If yours was near 150,
you handed `rlnorm()` a raw mean.

## 4. The choices

```{r}
n_per_group <- 60      # 120 surveys, split evenly
effect <- 1.5          # detect a 50% difference
```

Neither of these came from the data. The budget is the agency's, and the effect
size is a judgement about what difference would matter.

::: {.notes}
An even split is the right default when the two groups have the same spread.
:::

## 5. Simulate, analyze, repeat {.smaller}

```{r}
power_reef <- function(n_per_group, effect, nsim = 500,
                       meanlog = meanlog_unprot, sdlog = sdlog_unprot) {

  found <- numeric(nsim)

  for (i in 1:nsim) {

    sim_unprot <- rlnorm(n_per_group, meanlog = meanlog, sdlog = sdlog)
    sim_prot   <- rlnorm(n_per_group, meanlog = meanlog + log(effect),
                         sdlog = sdlog)

    found[i] <- t.test(log(sim_unprot), log(sim_prot))$p.value < 0.05

  }

  mean(found)

}

power_reef(n_per_group = 60, effect = 1.5)
```

::: {.notes}
The comparison is on the log scale, which is both where the data are symmetric
and where the effect is a constant shift. A t-test on raw abundance would also
run, and would have less power.
:::

## How many surveys do you need?

```{r}
#| echo: false
#| fig-height: 3.2
#| fig-width: 7
ns <- c(20, 40, 60, 100, 200, 400)
p50 <- sapply(ns, power_reef, effect = 1.5)
p23 <- sapply(ns, power_reef, effect = 1.23)
plot(ns, p50, type = "b", pch = 16, lwd = 2, ylim = c(0, 1), col = "firebrick",
     xlab = "surveys per group", ylab = "power")
lines(ns, p23, type = "b", pch = 16, lwd = 2, col = "steelblue")
abline(h = 0.8, lty = 2, col = "blue")
text(20, 0.85, "0.8: the usual target", col = "blue", cex = 0.9, adj = 0)
legend("right", lwd = 2, pch = 16, col = c("firebrick", "steelblue"), bty = "n",
       legend = c("50% difference", "23%: what these reefs show"))
```

```{r}
#| echo: false
knitr::kable(data.frame(surveys_per_group = ns,
                        power_50pct = round(p50, 2),
                        power_23pct = round(p23, 2)))
```

## What this tells the manager {.smaller}

- A **50% difference** is detectable with roughly **80 to 100 surveys per
  group**.
- The difference these reefs actually show, about **23%**, needs roughly
  **400 per group**.
- With the 120 surveys on offer, split evenly, power for the real difference is
  around **0.2**.

. . .

So: protection may well work, and this survey could not show it.
**Failing to detect is not evidence of no effect.**

::: {.notes}
That last line is the third warning from the lecture, met in a real case.

The useful advice to a manager is not "do the survey anyway". It is either find
more budget, pick a more sensitive response than total abundance, or accept that
the study answers a different question.
:::
