---
title: "The bestiary of functions"
subtitle: "PCB 6455 / BSC 4933: Biometry &nbsp;&nbsp;|&nbsp;&nbsp; Lecture 10"
author: "Dr. Brian Maitner"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    code-line-numbers: false
    fig-width: 7
    fig-height: 4.2
    fig-align: center
    html-math-method: mathml
    include-in-header:
      text: |
        <style>
        .type-it { border-left: 6px solid #2a7ab0; padding-left: 0.4em; }
        .type-it::before, .type-badge {
          content: "Type and run"; display: inline-block;
          background: #2a7ab0; color: white; font-weight: bold;
          font-size: 0.5em; line-height: 1.4; padding: 0 0.6em;
          border-radius: 4px; vertical-align: middle;
        }
        .type-badge { font-size: 0.7em; }
        </style>
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(emdbook)
data(ReedfrogSizepred)
data(ReedfrogFuncresp)
tadpoles <- ReedfrogSizepred
kills_by_density <- ReedfrogFuncresp

avonet <- read.csv("../data/Avonet/AVONET1_BirdLife.csv")

ricker <- function(x, a = 1, b = 1) a * x * exp(-b * x)
michaelis_menten <- function(x, a = 1, b = 1) a * x / (b + x)
power_ricker <- function(x, height, peak, alpha) height * (x / peak * exp(1 - x / peak))^alpha
hockey_stick <- function(x, a, s) ifelse(x < s, a * x, a * s)
```

## Today's agenda

- Reading a function by its shape
- The right curve depends on the question
- Modifying a function to fit
- Piecewise functions and derivatives in R (3.6)

. . .

[Type and run]{.type-badge} on a code block: type it into your script and run it.
You will need it later.

Code without the label is there to watch.

::: {.notes}
Say this out loud, as in Lecture 9. Some code on these slides is a worked example to follow on the
screen, and some is code they will need a few slides later. Only the second kind is
labeled, so if something is labeled and they skip it, a later exercise breaks.
:::

## Where we left off

```{r}
#| echo: false
plot(tadpoles$TBL, tadpoles$Kill,
     xlab = "Tadpole body length (mm)", ylab = "Number killed",
     pch = 16, cex = 1.4, xlim = c(0, 40), ylim = c(0, 6))
curve(ricker(x, a = 4 * (1/12) * exp(1), b = 1/12),
      add = TRUE, col = "red", lwd = 3)
```

The Ricker had the right kind of shape, but its right tail is too heavy.
Score: **87.5**.

::: {.notes}
The closing figure from Lecture 9. The eyeballed Ricker keeps predicting kills at
25 and 37 mm, where the data say zero, and that is most of its 87.5.

Today picks that thread straight back up. By the end, students will have fixed
this fit.
:::

## Two ways forward

1. **Try a different function**, one whose shape matches better.
2. **Modify the function you have**, so its shape changes.

. . .

Either way you need a vocabulary of shapes. That is what the **bestiary** is.

::: {.notes}
Bolker's own advice about the bestiary is to skim it on a first reading and use it
as a reference, not to memorize it. So today is about how to use it: how to find
a shape, and how to bend one that is close.

Both routes get used today, and the tadpoles get fixed by the second one.
:::

## Get set up {background-color="#f0f4f8"}

Run this so the rest of today works.

::: {.type-it}
```{r}
#| eval: false
library(emdbook)
data(ReedfrogSizepred)
data(ReedfrogFuncresp)
tadpoles <- ReedfrogSizepred
kills_by_density <- ReedfrogFuncresp

ricker <- function(x, a = 1, b = 1){
  a * x * exp(-b * x)
}

michaelis_menten <- function(x, a = 1, b = 1){
  a * x / (b + x)
}
```
:::

::: {.notes}
Two minutes. The two functions are the ones they wrote last lecture, so typing
them again is a quick rehearsal. Anyone who still has last lecture's script open
can run it from there.

ReedfrogFuncresp is new: a second tadpole experiment, used in a few slides.
:::

## Reading a function by its shape {.center}

## Read a function by its ends and its middle

Three questions to ask of any function:

1. What happens at the **left end**?
2. What happens at the **right end**?
3. What happens **in between**? Does it rise, fall, peak, level off, or switch?

. . .

Bolker's Table 3.1 sorts the whole bestiary by exactly these.

::: {.notes}
They already asked the first two of the Ricker last lecture, on the "what happens
at the ends" slide. This generalizes it into a way of reading any function.

The middle is where the shape names come from: saturating, S-shaped, humped, and
so on, which is the next slide.
:::

## A catalog of shapes

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 5
par(mfrow = c(2, 3), mar = c(2.5, 2.5, 3, 0.8), mgp = c(1.4, 0.5, 0))
curve(michaelis_menten(x, a = 1, b = 1), 0, 10, lwd = 3, col = "red", ylim = c(0, 1),
      xlab = "", ylab = "", main = "Saturating\nMichaelis-Menten")
curve(plogis(x, location = 5, scale = 1), 0, 10, lwd = 3, col = "red", ylim = c(0, 1),
      xlab = "", ylab = "", main = "S-shaped (sigmoid)\nlogistic")
curve(ricker(x, a = 1, b = 1), 0, 8, lwd = 3, col = "red",
      xlab = "", ylab = "", main = "Humped\nRicker")
curve(exp(-0.5 * x), 0, 10, lwd = 3, col = "red", ylim = c(0, 1),
      xlab = "", ylab = "", main = "Decreasing\nnegative exponential")
curve(x^2, 0, 3, lwd = 3, col = "red", ylim = c(0, 9),
      xlab = "", ylab = "", main = "Power law\na x^b")
curve(sqrt(x) * 3, 0, 3, lwd = 3, col = "blue", add = TRUE)
curve(hockey_stick(x, a = 1, s = 5), 0, 10, lwd = 3, col = "red", ylim = c(0, 6),
      xlab = "", ylab = "", main = "Switch\nhockey stick")
```

::: {.notes}
Six shapes cover most of what they will meet. Each panel is one representative;
the bestiary has several functions for most shapes.

On the power law panel, red has b > 1 (accelerating) and blue has 0 < b < 1
(decelerating). Both are a x^b.

Point out that they already know three of these from last lecture: saturating,
humped, and decreasing.
:::

## Your turn: which shape would you try first? {background-color="#f0f4f8" .smaller}

No code. For each, name a starting shape and one reason.

1. Seeds eaten by a mouse ~ seed density
2. Light level ~ depth in a lake
3. Bird body mass ~ leg length
4. Probability a tree has flowered ~ trunk diameter
5. Total offspring produced ~ population density
6. Food eaten by a filter feeder with a hard maximum ~ food concentration

Read `~` as "depends on": **response ~ predictor**, the y axis on the left and
the x axis on the right.

More than one answer can be defensible. The reason is the point.

::: {.notes}
Three to five minutes in pairs, then go around. Deliberately low stakes.

The tilde is R's formula notation, met already in Lecture 5 as
boxplot(log10(Mass) ~ Trophic.Level, data = avonet) and in the lm() calls in the
chapter 1 script. Worth naming it here, because it is how every model gets
written from chapter 9 on, and it is on the AVONET slide later today.

Answers to listen for:

1. Saturating. A mouse can only handle so many seeds; handling time, the Holling
   argument from last lecture.
2. Decreasing, negative exponential. Each meter of water absorbs a constant
   fraction of the light that reaches it. Bolker mentions the same law for light
   in plant canopies.
3. Power law. Bolker's own example.
4. S-shaped. Small trees almost never flower, big ones almost always do, with a
   switch in between.
5. Humped. The Ricker was built for this: more adults means more offspring, until
   crowding cuts fecundity.
6. Switch, hockey stick. Intake rises until it hits the limit and then flattens
   suddenly. Bolker calls this the Holling type I response.

If someone argues a sigmoid for 1, that is a real alternative (Holling type III),
and it is worth saying so.
:::

## The right curve depends on the question {.center}

## A second tadpole experiment

```{r}
#| echo: false
#| fig-height: 3.8
plot(kills_by_density$Initial, kills_by_density$Killed,
     xlab = "Number of tadpoles in the tank", ylab = "Number killed",
     pch = 16, cex = 1.4, xlim = c(0, 100), ylim = c(0, 40))
```

Same predators, same prey. This time the question is **how many** tadpoles, not
**how big**. What shape is this?

::: {.notes}
ReedfrogFuncresp, from the same research program as last lecture's size data.

Let them answer. It rises and seems to flatten: saturating. The biology is the
handling-time argument: with more tadpoles around, predators spend more of their
time eating and less searching, so kills cannot keep rising in proportion.
:::

## Michaelis-Menten fits this one

```{r}
#| fig-height: 3.2
plot(kills_by_density$Initial, kills_by_density$Killed,
     xlim = c(0, 100), ylim = c(0, 40),
     xlab = "Number of tadpoles in the tank", ylab = "Number killed")
curve(michaelis_menten(x, a = 60, b = 100), add = TRUE, col = "red", lwd = 2)
```

The curve that failed on **size** in Lecture 9 fits **density**.

::: {.notes}
This is the payoff of the section title. Last lecture Michaelis-Menten was set
aside because size data rise and then fall. Here the pattern really does saturate,
for a mechanistic reason. The function did not change; the question did.

It also closes the loop on the two formulations from last lecture: this is a
functional response, so the mechanistic Holling form, with attack rate and
handling time, is a natural way to write it.
:::

## Your turn: how high does it go? {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
a_try <- 60                    # the asymptote
b_try <- 100                   # the half-saturation point

plot(kills_by_density$Initial, kills_by_density$Killed,
     xlim = c(0, 100), ylim = c(0, 40))
curve(michaelis_menten(x, a = a_try, b = b_try), add = TRUE, col = "red", lwd = 2)

sum((kills_by_density$Killed -
     michaelis_menten(kills_by_density$Initial, a_try, b_try))^2)
```
:::

Starting score: **404**. Now try `a_try <- 150` and `b_try <- 390`.

What does that tell you about the maximum number of kills?

::: {.notes}
About 5 minutes. The second pair scores about 376, nearly the same. With b adjusted
to match, any asymptote from about 60 to 300 fits these data about equally well.

The lesson: the data never get near the asymptote, so they cannot tell you where it
is. Bolker warns about exactly this for Michaelis-Menten: it approaches its
asymptote very slowly. That says more about what data you would need than about
the function, and it is worth remembering before reporting a maximum from a fit.

Cuttable if time is short; the previous slide makes the main point.
:::

## Modifying a function to fit {.center}

## Four ways to change a function

Start from any $f(x)$:

| Change | Math | Effect |
|---|---|---|
| **Scale** | $a \cdot f(x)$ | taller or shorter |
| **Stretch** | $f(x / b)$ | wider or narrower |
| **Shift** | $f(x - c)$ | slides right |
| **Bend** | $f(x)^{\alpha}$ | changes the shape itself |

::: {.notes}
Bolker covers the first three under shifting and scaling in 3.3, which they read
for last lecture, and the fourth under power laws in the bestiary.

The first three move or resize a curve without changing its basic shape. The power
is different: it changes the shape, which is what the tadpoles need.

They have already seen the first two in action. In Michaelis-Menten, a scales and
b stretches: that is why a is the height and b is where it gets halfway.
:::

## What each one does

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
base <- function(x) x / (1 + x)
par(mfrow = c(2, 2), mar = c(2.5, 2.5, 2.5, 0.8), mgp = c(1.4, 0.5, 0))
curve(base(x), 0, 10, lty = 2, lwd = 2, col = "grey50", ylim = c(0, 2),
      xlab = "", ylab = "", main = "Scale:  2 f(x)")
curve(2 * base(x), add = TRUE, col = "red", lwd = 3)
curve(base(x), 0, 10, lty = 2, lwd = 2, col = "grey50", ylim = c(0, 1),
      xlab = "", ylab = "", main = "Stretch:  f(x / 3)")
curve(base(x / 3), add = TRUE, col = "red", lwd = 3)
curve(base(x), 0, 10, lty = 2, lwd = 2, col = "grey50", ylim = c(0, 1),
      xlab = "", ylab = "", main = "Shift:  f(x - 2)")
curve(base(pmax(x - 2, 0)), add = TRUE, col = "red", lwd = 3)
curve(base(x), 0, 10, lty = 2, lwd = 2, col = "grey50", ylim = c(0, 1),
      xlab = "", ylab = "", main = "Bend:  f(x) ^ 3")
curve(base(x)^3, add = TRUE, col = "red", lwd = 3)
```

Gray dashed: $f(x) = x / (1 + x)$. Red: the changed version.

::: {.notes}
Point at the bend panel. Cubing a saturating curve turns it S-shaped: flat near
zero, then rising. The first three panels are the same shape moved around; the
fourth is a different shape.

On the shift panel the curve is held at zero before x = 2. Without that, x/(1+x)
shifted right hits a division by zero, which Bolker warns about.
:::

## Power laws

$$f(x) = a\,x^{b}$$

- $a$: scale, the value at $x = 1$
- $b$: the exponent. $b > 1$ accelerates, $0 < b < 1$ decelerates, $b < 0$ decreases

```{r}
#| eval: false
power_law <- function(x, a = 1, b = 1){
  a * x^b
}
```

On log-log axes, a power law is a **straight line** with slope $b$.

::: {.notes}
Callback to Lecture 5: the log-log axes that turned the mass versus wing length
cloud into a line. That line was a power law, and its slope was the exponent.

Power laws describe allometry, metabolic rates, and species-area curves, among
others.
:::

## A power law in AVONET

```{r}
#| echo: false
#| fig-height: 3.9
fit <- lm(log10(Mass) ~ log10(Tarsus.Length), data = avonet)
plot(log10(avonet$Tarsus.Length), log10(avonet$Mass),
     pch = 16, cex = 0.3, col = rgb(0, 0, 0, 0.2),
     xlab = "log10 tarsus length (mm)", ylab = "log10 body mass (g)")
abline(fit, col = "red", lwd = 3)
```

Slope about **2.2**: body mass rises roughly as leg length to the power 2.2.

::: {.notes}
Bolker's own example of a power law is bird mass against tarsus length, and AVONET
has both columns.

If you want the number: lm(log10(Mass) ~ log10(Tarsus.Length), data = avonet)
gives a slope of 2.17.

Worth a caution, and Bolker makes it: if birds were scaled-up copies of each
other, mass would go as length cubed, a slope of 3. The slope is 2.2. It is
tempting to read a mechanism off that number, but many different processes produce
power laws, so an exponent on its own is weak evidence for any one of them.
:::

## Fixing the tadpoles {.center}

## The Ricker's hump is too wide

A power **bends** a curve's shape.

Raise the Ricker to a power greater than 1 and the hump gets **narrower**, with
lighter tails.

. . .

That is exactly what the tadpole data need.

::: {.notes}
Tie back to the bend panel: a power changed the shape of a saturating curve. Here
it does the same to a humped one.

Bolker's Figure 3.1 shows this fix: the power-Ricker, from Persson and colleagues,
fits the tadpole size data far better than the plain Ricker.
:::

## The only change is an exponent {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Ricker**, as you wrote it last time

$$f(x) = a\,x\,e^{-b x}$$

the same curve, written with the peak's height $h$ and position $p$

$$f(x) = h \left( \frac{x}{p}\, e^{\,1 - x/p} \right)$$
:::

::: {.column width="50%"}
**Power-Ricker**

$$f(x) = h \left( \frac{x}{p}\, e^{\,1 - x/p} \right)^{\alpha}$$

Everything in the brackets is the Ricker.

The only new thing is the $\alpha$.
:::

::::

::: {.notes}
Two moves, so take them one at a time.

First, down the left column: the same curve rewritten. Both lines are the Ricker;
they just carry different parameters. They match when a = h e / p and b = 1 / p,
so last lecture's eyeballed fit, peak 4 killed at 12 mm, is h = 4 and p = 12. The
reason for rewriting is the one from last lecture's two versions of
Michaelis-Menten: choose the parameters you can read off a plot.

Then across: the brackets are identical, and the exponent is the whole
modification. This is the "bend" row of the four-ways table, applied to a curve
they already have.

The shapes those equations make are on the next slide.
:::

## Side by side

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.8
par(mfrow = c(1, 2), mar = c(3.2, 3.2, 2.6, 0.8), mgp = c(1.9, 0.6, 0))
curve(power_ricker(x, 4, 12, 1), from = 0, to = 40, ylim = c(0, 4.4),
      col = "red", lwd = 3, main = "Ricker", xlab = "x", ylab = "f(x)")
abline(v = 12, lty = 3, col = "grey50")
curve(power_ricker(x, 4, 12, 1), from = 0, to = 40, ylim = c(0, 4.4),
      col = "grey60", lwd = 2, lty = 2,
      main = "Power-Ricker, raised to the power 3", xlab = "x", ylab = "f(x)")
curve(power_ricker(x, 4, 12, 3), add = TRUE, col = "red", lwd = 3)
abline(v = 12, lty = 3, col = "grey50")
```

Same peak, same height. The power **narrows the hump** and **pulls the right
tail down**.

::: {.notes}
The picture behind the claim on the last slide, before any equation.

Gray dashed on the right is the plain Ricker from the left panel, so the change
is visible without looking back and forth between slides.

Two things to point at, because they are what the tadpoles need: the right tail
reaches zero far sooner, and the rise on the left is steeper and starts later.
The peak does not move, which is the whole reason this version is convenient.

The function that drew this arrives on the next slide. Nobody needs to type it
yet.
:::

## The power-Ricker, in math and in R {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Math**

$$f(x) = h \left( \frac{x}{p}\, e^{\,1 - x/p} \right)^{\alpha}$$

- $h$: the height of the peak
- $p$: where the peak is
- $\alpha$: how narrow the hump is ($\alpha > 0$)

With $\alpha = 1$ it **is** the Ricker.
:::

::: {.column width="50%"}
**R**

```{r}
#| eval: false
power_ricker <- function(x, height, peak, alpha){
  height * (x / peak * exp(1 - x / peak))^alpha
}
```

- the Ricker part is inside the brackets
- `^alpha` is the bend
:::

::::

::: {.notes}
Written with the peak's height and position as parameters, which are exactly the
two numbers read off the plot last lecture: 4 killed, at 12 mm. Same idea as last
lecture's two ways of writing Michaelis-Menten: choose the parameters you can see.

With alpha = 1 this is algebraically the Ricker with a = h e / p and b = 1 / p, so
height 4 and peak 12 reproduces last lecture's eyeballed fit exactly. The students
will confirm that in a moment, because it scores 87.5 again.
:::

## Your turn: fix the fit {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
power_ricker <- function(x, height, peak, alpha){
  height * (x / peak * exp(1 - x / peak))^alpha
}

alpha_try <- 1                 # change this

plot(tadpoles$TBL, tadpoles$Kill, xlim = c(0, 40), ylim = c(0, 6))
curve(power_ricker(x, height = 4, peak = 12, alpha = alpha_try),
      add = TRUE, col = "red", lwd = 2)

sum((tadpoles$Kill - power_ricker(tadpoles$TBL, 4, 12, alpha_try))^2)
```
:::

With `alpha_try <- 1` you should get **87.5**, last lecture's score.
The best any Ricker can do is **24.9**. Can you beat it?

::: {.notes}
About 8 minutes. The check at alpha = 1 matters: it confirms their function is
typed correctly and shows the power-Ricker contains the Ricker.

Scores to expect with height 4 and peak 12: alpha 2 gives about 58, alpha 4 about
34, alpha 6 about 24, which beats the best plain Ricker. Anyone who keeps going
will find it keeps improving, which sets up the slide after next.
:::

## What $\alpha$ does

```{r}
#| echo: false
plot(tadpoles$TBL, tadpoles$Kill,
     xlab = "Tadpole body length (mm)", ylab = "Number killed",
     pch = 16, cex = 1.4, xlim = c(0, 40), ylim = c(0, 6))
curve(power_ricker(x, 4, 12, 1), add = TRUE, col = "grey50", lwd = 2, lty = 2)
curve(power_ricker(x, 4, 12, 2), add = TRUE, col = "orange", lwd = 2)
curve(power_ricker(x, 4, 12, 6), add = TRUE, col = "red", lwd = 3)
legend("topright", bty = "n", lwd = c(2, 2, 3), lty = c(2, 1, 1),
       col = c("grey50", "orange", "red"),
       legend = c("alpha = 1 (the Ricker), SS 87.5", "alpha = 2, SS 58.4",
                  "alpha = 6, SS 24.0"))
```

Same peak, same height. Bigger $\alpha$, narrower hump.

::: {.notes}
The peak stays at 12 mm and 4 killed for every alpha, because those are their own
parameters. Only the width changes, which is why this parameterization is so easy
to use.
:::

## How far should you go?

```{r}
#| echo: false
#| fig-height: 3.8
plot(tadpoles$TBL, tadpoles$Kill,
     xlab = "Tadpole body length (mm)", ylab = "Number killed",
     pch = 16, cex = 1.4, xlim = c(0, 40), ylim = c(0, 6))
curve(power_ricker(x, 4.002, 11.939, 38.075), add = TRUE, col = "red", lwd = 3,
      n = 1001)
```

The best fit has $\alpha = 38$ and a score of **5**. **Do you believe it?**

::: {.notes}
Let the room argue. The spike fits the data very well, but look at what it claims
where there are no data: 4 kills at 12 mm, about 1.4 at 15 mm, and 0.1 at 18 mm.
Nobody measured tadpoles between 12 and 21 mm. The data cannot tell a sharp spike
from a gentler hump in that gap, so the curve is filling it with an assumption.

The general lesson: a more flexible function can always fit better, and a better
score does not mean a better description. Bolker notes that exponents like this
are hard to estimate from small data sets. Later chapters give tools, like
likelihood-based model comparison, for deciding when extra flexibility is earned.
Callback to Lecture 9's first slide: fit it, and then ask whether you believe it.
:::

## Piecewise functions and derivatives in R {.center}

## Piecewise functions with `ifelse()` {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Math**: a hockey stick

$$f(x) = \begin{cases} a\,x & \text{if } x < s \\ a\,s & \text{if } x \ge s \end{cases}$$

- $a$: the slope before the switch
- $s$: where it switches
:::

::: {.column width="50%"}
**R**

::: {.type-it}
```{r}
#| eval: false
hockey_stick <- function(x, a, s){
  ifelse(x < s, a * x, a * s)
}
```
:::

`ifelse(condition, if true, if false)`
:::

::::

```{r}
#| echo: false
#| fig-height: 2.6
#| fig-width: 6
par(mar = c(3, 3, 0.5, 0.5), mgp = c(1.6, 0.5, 0))
curve(hockey_stick(x, a = 2, s = 5), from = 0, to = 10, lwd = 3, col = "red",
      xlab = "x", ylab = "f(x)")
```

::: {.notes}
Section 3.6.2 of the R supplement.

The biology: a filter feeder whose intake rises with food concentration until it
physically cannot process any more, then stops rising. Bolker notes this is the
Holling type I functional response. Piecewise functions make sense when there is a
genuine switch point.

ifelse works on a whole vector at once, which is why it plays well with curve().
:::

## Your turn: a threshold {background-color="#f0f4f8"}

A threshold is flat, jumps at $s$, then is flat again:
$f(x) = a_1$ if $x < s$, otherwise $a_2$.

```{r}
#| echo: false
#| fig-height: 2.2
#| fig-width: 6
par(mar = c(3.2, 3.2, 0.4, 0.8), mgp = c(1.9, 0.6, 0))
plot(NULL, xlim = c(0, 10), ylim = c(0, 4), xlab = "x", ylab = "f(x)")
abline(v = 5, lty = 3, col = "grey50")
segments(0, 1, 5, 1, col = "red", lwd = 3)
segments(5, 3, 10, 3, col = "red", lwd = 3)
text(2.5, 1.45, expression(a[1]), cex = 1.4)
text(7.5, 3.45, expression(a[2]), cex = 1.4)
text(5.3, 0.25, "s", cex = 1.4, adj = 0)
```

::: {.type-it}
```{r}
#| eval: false
threshold <- function(x, a1, a2, s){

}

curve(threshold(x, a1 = 1, a2 = 3, s = 5), from = 0, to = 10)
```
:::

::: {.notes}
About 4 minutes. The answer is ifelse(x < s, a1, a2), one line, and it is a direct
adaptation of the hockey stick on the previous slide.

The figure is the target: a1 = 1 before the switch, a2 = 3 after it, s = 5, which
are the numbers in the curve() call below it. Their plot should look like this.
It shows the shape, not the code, so the exercise is still theirs.

If the plot shows a sloped line at the jump rather than a vertical step, that is
just curve() drawing a line between neighboring points; add n = 1001 to make the
step look sharper.
:::

## R can do the calculus

Last lecture the book told us the Ricker peaks at $x = 1/b$. Check it with `D()`:

::: {.type-it}
```{r}
d_ricker <- D(expression(a * x * exp(-b * x)), "x")
d_ricker

eval(d_ricker, list(a = 1, b = 0.25, x = 1 / 0.25))
eval(d_ricker, list(a = 2, b = 0.25, x = 0))
```
:::

Slope **0** at $x = 1/b$: that is the peak. Slope **$a$** at $x = 0$: the initial slope.

::: {.notes}
Section 3.6.3 of the R supplement. D() takes a derivative symbolically; eval() plugs
in numbers.

A peak is where the slope is zero, so a zero derivative at x = 1/b confirms last
lecture's claim without doing the calculus by hand. The second line confirms that a
is the initial slope, the other parameter meaning from last lecture.

Worth saying that R does not simplify its answer, which is why it looks messy.
:::

## Remainder of class

Complete **3.6**, the R supplement:

- 3.6.1 Plotting functions (most of today)
- 3.6.2 Piecewise functions with `ifelse()`
- 3.6.3 Derivatives with `D()` and `deriv()`

Code: `R_scripts/Chapter_3_examples.R`

::: {.notes}
Matches the syllabus row for Lecture 10: "Read 3.4 to 3.5; Complete 3.6".
:::

## Before next class

Read Bolker **4.1 to 4.2**, and **4.4**

Next time: probability and stochastic distributions
