---
title: "Deterministic functions"
subtitle: "PCB 6455 / BSC 4933: Biometry &nbsp;&nbsp;|&nbsp;&nbsp; Lecture 9"
author: "Dr. Brian Maitner"
format:
  revealjs:
    theme: simple
    slide-number: true
    embed-resources: true
    code-line-numbers: false
    fig-width: 7
    fig-height: 4.2
    fig-align: center
    html-math-method: mathml
    include-in-header:
      text: |
        <style>
        .type-it { border-left: 6px solid #2a7ab0; padding-left: 0.4em; }
        .type-it::before, .type-badge {
          content: "Type and run"; display: inline-block;
          background: #2a7ab0; color: white; font-weight: bold;
          font-size: 0.5em; line-height: 1.4; padding: 0 0.6em;
          border-radius: 4px; vertical-align: middle;
        }
        .type-badge { font-size: 0.7em; }
        </style>
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(emdbook)
data(ReedfrogSizepred)
tadpoles <- ReedfrogSizepred

ricker <- function(x, a = 1, b = 1) a * x * exp(-b * x)
negexp <- function(x, a = 1, b = 1) a * exp(-b * x)
michaelis_menten <- function(x, a, b) a * x / (b + x)
holling_type2 <- function(x, alpha, h) alpha * x / (1 + alpha * h * x)
```

## Today's agenda

- What a function is, and why you would write your own
- Finding a curve with the right shape for real data
- What the parameters do
- Fitting it by eye

. . .

[Type and run]{.type-badge} on a code block: type it into your script and run it.
You will need it later.

Code without the label is there to watch.

::: {.notes}
Say this out loud. Some code on these slides is a worked example to follow on the
screen, and some is code they will need a few slides later. Only the second kind is
labeled, so if something is labeled and they skip it, a later exercise breaks.
:::

## Where we are

**Chapter 2:** *describe* the pattern.
Histograms, boxplots, scatterplots.

**Chapter 3:** *name* the pattern.
Write down a function that has that shape.

**Chapter 4 onward:** fit it, and ask whether you believe it.

::: {.notes}
Bolker's opening line is that chapter 2 gives "only qualitative descriptions",
and that you avoid committing too early on purpose. Now is when you commit.

Be concrete about why: a curve with named parameters lets you say how fast, how
big, where the peak is, and compare those numbers between sites, years or
species. A scatterplot cannot do that.
:::

## You have been using functions all semester

```{r}
#| eval: false
read.csv()      mean()      hist()
plot()          filter()    summarise()
ggplot()        log10()     install.packages()
```

Every one works the same way:

**inputs go in, something happens, a result comes back.**

::: {.notes}
Ask the room to name more before moving on. They have used dozens: str(),
table(), which.max(), curve(), cor(), lm(). The point is that none of this is
new, it just has not had a name.

The block is shown as code for the formatting, not to run.
:::

## Anatomy of a function call

```{r}
mean(c(2, 4, NA))
mean(c(2, 4, NA), na.rm = TRUE)
```

- `mean` is the **name**
- `c(2, 4, NA)` and `na.rm = TRUE` are the **arguments**
- `na.rm` has a **default**, `FALSE`, used unless you say otherwise

::: {.notes}
Callback to Lecture 4 and to Quiz 2, so the na.rm behavior should be familiar.
What is new is the vocabulary: name, arguments, default.

?mean shows the arguments and their defaults. Worth saying that every help page
they have opened has a Usage section that is exactly this.

Defaults come back later today when they write their own.
:::

## Functions are just code someone wrote

Type a function's name without the brackets:

```{r}
sd
```

`sd()` is `sqrt()` of `var()`, with a default for `na.rm`.

You can write these too.

::: {.notes}
This is the reveal. A built-in function is not magic, it is a few lines of R that
someone saved under a name, built from other functions they already know.

Point at function (x, na.rm = FALSE): that is the same syntax they will write in
a few slides, default and all.

Ignore the bytecode and environment lines; they say where R stored it.
:::

## Why write your own?

- **Stop copying and pasting.** Lecture 6's four near-identical blocks: write it
  once instead.
- **Fix mistakes in one place.** One copy means one fix.
- **Give an idea a name.** `standard_error(x)` says what it does.
- **An equation is a function.** $f(x)$ in math, `function(x)` in R.
  That is this chapter.

::: {.notes}
The first two are programming reasons and they have felt both: the four-panel
copy-paste in Lecture 6, and changing ylim in four places.

The last one is the reason for today. A curve like the Ricker is a function in
the math sense, and writing it as an R function is what lets you draw it, change
its parameters, and later fit it. From chapter 6 on, fitting a model means handing
R your model written as a function, so this is not a detour.
:::

## Writing your own

```{r}
standard_error <- function(x){
  sd(x) / sqrt(length(x))
}

standard_error(c(2, 4, 6, 8))
```

- `standard_error` is the name you choose
- `function(x)` says what arguments it takes
- the body in `{ }` does the work; the last line is the result
- `x` is a **placeholder**: whatever you pass in becomes `x`

::: {.notes}
A genuinely useful first example, because R has no built-in standard error of the
mean. There is a function called stderr(), but it is the stream R writes error
messages to, and stderr(c(2, 4)) errors. Good for a laugh, and a real trap.

The placeholder point matters for the rest of the lecture. The x inside the
function is not a variable called x in their workspace; it is whatever they hand
in. That is why ricker(tadpoles$TBL, ...) works later, and it is related to why
curve() insists on the letter x.
:::

## Your turn {background-color="#f0f4f8"}

To convert Celsius to Fahrenheit: $F = \dfrac{9}{5}\,C + 32$

::: {.type-it}
```{r}
#| eval: false
# Here is the other direction:
f_to_c <- function(temp_f){
  (temp_f - 32) * 5 / 9
}

f_to_c(98.6)

# Write this one, from the math above:
c_to_f <- function(temp_c){

}

c_to_f(c(0, 37, 100))     # should give 32, 98.6, 212
```
:::

::: {.notes}
Short pause, about 5 minutes. Two things are being rehearsed at once: writing a
function, and turning an equation into R, which is exactly what they do with the
Ricker shortly.

The check on the last line matters. They can see immediately whether it worked,
without asking.

Cuttable if time is short. The anatomy slide covers the syntax and the Ricker
pause later exercises it.
:::

## Now let's use them

You can write a function.

So far, functions that convert one number into another.

Next: functions that **describe a pattern in data**.

::: {.notes}
Transition slide, thirty seconds. The Celsius converter and the standard error were
about how functions work. From here on, the functions are curves, and the question
is which one matches the data.
:::

## The question

```{r}
#| echo: false
plot(tadpoles$TBL, tadpoles$Kill,
     xlab = "Tadpole body length (mm)", ylab = "Number killed",
     pch = 16, cex = 1.4, xlim = c(0, 40), ylim = c(0, 6))
```

**What curve would you draw through this?**

::: {.notes}
Bolker's Figure 3.1, data in emdbook as ReedfrogSizepred. Only 15 points.

Let them answer before moving on. Every curve for the rest of the lecture gets
held up against this plot, so it is worth a minute.

They met tadpoles in Lecture 3, in the frogs and tadpoles regression.
:::

## What shape do we need? {.smaller}

Load the data and plot it yourself:

::: {.type-it}
```{r}
#| eval: false
library(emdbook)
data(ReedfrogSizepred)
tadpoles <- ReedfrogSizepred
plot(tadpoles$TBL, tadpoles$Kill)
```
:::

Reading it off the plot:

- It **rises** from the smallest tadpoles to about 12 mm
- It **peaks** around 12 mm
- It **falls back** toward zero for the biggest tadpoles

We need a function that can do all three.

::: {.notes}
Have everyone run the four lines now. The tadpole data are used on the next several
slides and in the last exercise, and it is much quicker to sort out a missing
package here than halfway through.

This list is the test every candidate faces today. Point back at it each time a
curve fails.

The biology behind each part, worth getting the room to supply: small tadpoles
hide or are ignored, large ones escape or are too big to swallow, and something
in between is the worst size to be.
:::

## Two kinds of description

**Phenomenological:** a curve that fits.
Chosen because it has the right shape.

**Mechanistic:** a curve derived from a theory about the process.
Parameters mean something.

. . .

The same equation can be either, depending on why you chose it.

::: {.notes}
Callback to Lecture 2. Bolker makes it explicitly in 3.1: "the same function can
be either phenomenological or mechanistic depending on context."

For the tadpoles we will mostly shop by shape, which is the phenomenological
route. But keep an eye on what each parameter means, because that is what lets
you read parameter values off the plot at the end.
:::

## First try: a curve you already know

The Holling type II from Lecture 2 **rises from zero**.

That is the first item on our list. Could it work?

::: {.notes}
The first reason to try any curve: it already has part of the shape you need. And
starting with one they know is the natural move.

Before testing it against the data, a short detour that matters: this curve can
be written two ways, and which one you write depends on why you are using it.
:::

## One curve, two ways to write it {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Mechanistic**\
Holling type II functional response

$$f(x) = \frac{\alpha\,x}{1 + \alpha\,h\,x}$$

- $x$: prey density
- $\alpha$: attack rate
- $h$: handling time

Written down from a theory of how predators forage.
:::

::: {.column width="50%"}
**Phenomenological**\
Michaelis-Menten form

$$f(x) = \frac{a\,x}{b + x}$$

- $x$: prey density
- $a$: asymptote, the maximum
- $b$: half-saturation, where $f(x) = a/2$

Written down because it has the right shape.
:::

::::

. . .

**The same curve**, when $a = \dfrac{1}{h}$ and $b = \dfrac{1}{\alpha\,h}$.

::: {.notes}
This is the concrete version of the "two kinds of description" slide, and Bolker
makes the point in chapter 1: the same function can be used mechanistically
"because you expected predators to attack prey at a constant rate and be
constrained by handling time", or phenomenologically "because you wanted a
function that started at zero, was initially linear, and leveled off".

What changes is not the curve but which parameters you write down, and that
choice follows from why you are using it:

- Mechanistic form when the parameters are the question. Do handling times differ
  between prey types? Does attack rate rise with seed size? You can only ask that
  if alpha and h are in the equation. Also safer to extrapolate.
- Michaelis-Menten form when you want parameters you can read off a plot. The
  maximum and the half-way point are both visible on a graph; attack rate and
  handling time are not.

How they connect, worth saying because Lecture 2 described a and b as attack rate
and handling time: related, but not the same. The maximum a is 1/h, so it is set
by handling time. The attack rate alpha is the initial slope, which in the second
form is a/b. As x gets large the first form tends to alpha x / (alpha h x) = 1/h,
which is a.
:::

## Same curve in R

You wrote the second one in Lecture 2, as `holling2()`.

```{r}
holling_type2 <- function(x, alpha, h){
  alpha * x / (1 + alpha * h * x)
}

michaelis_menten <- function(x, a, b){
  a * x / (b + x)
}

holling_type2(x = c(1, 5, 50), alpha = 2, h = 0.5)
michaelis_menten(x = c(1, 5, 50), a = 1 / 0.5, b = 1 / (2 * 0.5))
```

Identical numbers. **Which one you write depends on why you are using it.**

::: {.notes}
Let the two output lines sit next to each other for a moment. The same numbers
from two differently written functions is the most convincing way to show these
are one curve.

Point at the translation: the Greek letter becomes alpha, because you cannot type
it, and every multiplication needs a *.

The names are deliberate. In Lecture 2 this was holling2(), but written with a and
b it is the Michaelis-Menten form, so it gets that name here. Naming each function
for the form it takes is the "give an idea a name" reason from earlier, and it
keeps the two versions from being confused.
:::

## Math to R {.smaller}

| Math | R | |
|---|---|---|
| $a\,x$ | `a * x` | every multiplication needs `*` |
| $\dfrac{a}{b}$ | `a / b` | |
| $x^2$ | `x^2` | |
| $e^{-b x}$ | `exp(-b * x)` | a function, not a letter |
| $e$ | `exp(1)` | there is no `e` in R |
| $\ln(2)$ | `log(2)` | **`log()` is the natural log** |
| $\log_{10}(x)$ | `log10(x)` | |
| $\sum_i y_i$ | `sum(y)` | |

::: {.notes}
A reference for the rest of the lecture. Two rows are traps worth pointing at.

log(): in math ln is the natural log and log often means base 10. In R, log() is
the natural log and base 10 is log10(). Students who have used log10() since
Lecture 5 will reach for it when the book writes ln. This comes up with the
half-life.

e: typing e gives "object 'e' not found". The number e is exp(1). The exponential
row comes up in the next exercise, and exp(1) in the eyeball fit.

Cuttable if time is short; the pairings on the function slides make the same
points one at a time.
:::

## Defaults

::: {.type-it}
```{r}
michaelis_menten <- function(x, a = 1, b = 1){
  a * x / (b + x)
}
```
:::

`a = 1` sets a **default**. Anything you leave out uses it.

::: {.type-it}
```{r}
michaelis_menten(x = 2)
michaelis_menten(x = 2, a = 3)
```
:::

::: {.notes}
They met defaults earlier today, as na.rm = FALSE in mean() and sd(). Here they
set their own, with the same syntax they saw when they printed sd.

Have everyone type and run this definition. The next two slides draw it, and the
negative exponential exercise assumes they already have it. The holling_type2()
and michaelis_menten() on the previous slide were only for watching.
:::

## Drawing it

::: {.type-it}
```{r}
curve(michaelis_menten(x, a = 4, b = 5), from = 0, to = 40)
```
:::

::: {.notes}
curve() insists the variable is called x. That is a quirk of curve(), not a rule
about functions.

A trap someone will hit later: curve(x, ...) on its own fails with 'could not
find function "x"', because a bare name is read as a function. Write
curve(1 * x, ...). It comes up when we put two curves together.
:::

## Try it against the data

::: {.type-it}
```{r}
#| fig-height: 3.0
plot(tadpoles$TBL, tadpoles$Kill, xlim = c(0, 40), ylim = c(0, 6),
     xlab = "Tadpole body length (mm)", ylab = "Number killed")
curve(michaelis_menten(x, a = 4, b = 5), add = TRUE, col = "red", lwd = 2)
```
:::

Rises: **yes**. Peaks and falls back: **no**. It levels off and stays there.

::: {.notes}
add = TRUE draws the curve on top of the existing plot. Forgetting it silently
replaces the data with the curve, which someone will do.

The verdict, against the list: right start, wrong ending. And no choice of a and
b fixes it. A Michaelis-Menten curve can only rise and level off; it has no way to
come back down. So set it aside, and ask what it was missing: something that
falls.

If anyone asks what the "best" Michaelis-Menten fit would be: a fitting routine
only gets close by making b negative, a half-saturation point below zero, which
means nothing. That is a curve being forced into a shape it cannot make.
:::

## Second try: something that falls

Large tadpoles escape, so predation should **fall** with size.

The simplest curve that falls: the **negative exponential**.

::: {.notes}
Setting a curve aside for what it cannot do tells you what to look for next. The
Holling curve could not fall, so find one that does.

Why this one: it is the simplest decreasing curve there is, with just two
parameters. That also makes it the best place to learn what parameters do, which
is the next few slides.
:::

## Your turn {background-color="#f0f4f8" .smaller}

The negative exponential: $f(x) = a\,e^{-b x}$. Write it in R.

::: {.type-it}
```{r}
#| eval: false
# You typed this one in earlier:
michaelis_menten <- function(x, a = 1, b = 1){
  a * x / (b + x)
}

# Make this one, from the math above:
negexp <- function(x, a = 1, b = 1){

}

curve(negexp(x), from = 0, to = 7)

# Compare the two shapes: put the Michaelis-Menten curve on top
curve(michaelis_menten(x, a = 2, b = 1), add = TRUE, col = "red")
```
:::

::: {.notes}
First pause, about 8 minutes. The slide gives the math and they produce the R.
The Michaelis-Menten function is there as a model for the structure; the new
piece is the exponential, which is on the Math to R table as exp(-b * x).

The last step is where to circulate. The two curves have different ranges, so the
Michaelis-Menten curve will run off the top unless they add ylim. That is a real
problem with overlaying curves and worth them hitting. It also puts the two shapes
side by side: one rises, one falls.
:::

## The negative exponential, in math and in R

:::: {.columns}

::: {.column width="50%"}
**Math**

$$f(x) = a\,e^{-b x}$$

- $x$: the input, e.g. time
- $a$: the value at $x = 0$
- $b$: the decay rate ($b > 0$)
:::

::: {.column width="50%"}
**R**

```{r}
#| eval: false
negexp <- function(x, a = 1, b = 1){
  a * exp(-b * x)
}
```

- $e^{-b x}$ becomes `exp(-b * x)`
:::

::::

::: {.notes}
The answer to the pause. Check a few people got exactly this.

Why b > 0: that is what makes it fall. With b negative the same equation grows
without limit, and with b = 0 it is a flat line at a. The name "negative
exponential" assumes the positive case. The common mistakes
are writing e^(-b*x), which fails because there is no e, and exp(-bx), which fails
because bx is read as a single variable name.
:::

## Your turn {background-color="#f0f4f8" .smaller}

The last slide says what `a` and `b` do. **Check it**, changing only the numbers.

::: {.type-it}
```{r}
#| eval: false
# Round 1: three values of a, same b
curve(negexp(x, a = 1, b = 1), from = 0, to = 7, ylim = c(0, 3))
curve(negexp(x, a = 1, b = 1), add = TRUE, col = "red")
curve(negexp(x, a = 1, b = 1), add = TRUE, col = "blue")

# Round 2: same a, three values of b
curve(negexp(x, a = 1, b = 1), from = 0, to = 7, ylim = c(0, 3))
curve(negexp(x, a = 1, b = 1), add = TRUE, col = "red")
curve(negexp(x, a = 1, b = 1), add = TRUE, col = "blue")
```
:::

In your own words: what does `a` seem to control? What does `b` seem to control?

::: {.notes}
Second pause, about 7 minutes. All they change is numbers, which is exactly the
concept: same function, different parameters.

This comes before the two figure slides on purpose. Those slides are the answer:
show them once people have tried it, and let the room compare its own plots to
them.

Answers to listen for: a sets the height, the value the curve starts from, and
stretches it vertically. b sets how fast it falls, which stretches or squeezes the
curve horizontally. Let people describe b in their own terms; "how quickly it drops
off" and "how spread out it is" are both right.

Worth keeping in mind for later: in the Ricker, b controls where the peak sits,
so the horizontal effect of b becomes something you can point at.
:::

## What does `a` do?

```{r}
#| echo: false
curve(negexp(x, a = 1, b = 1), from = 0, to = 7, ylim = c(0, 3),
      xlab = "x", ylab = "y", lwd = 2)
curve(negexp(x, a = 2, b = 1), add = TRUE, col = "red",  lwd = 2)
curve(negexp(x, a = 3, b = 1), add = TRUE, col = "blue", lwd = 2)
legend("topright", lwd = 2, bty = "n", col = c("black", "red", "blue"),
       legend = c("a = 1", "a = 2", "a = 3"))
```

`a` is the value at x = 0.

::: {.notes}
Same shape, taller. a is a scale parameter: it stretches vertically and changes
nothing else.
:::

## What does `b` do?

```{r}
#| echo: false
curve(negexp(x, a = 1, b = 0.5), from = 0, to = 7, ylim = c(0, 1.1),
      xlab = "x", ylab = "y", lwd = 2)
curve(negexp(x, a = 1, b = 1), add = TRUE, col = "red",  lwd = 2)
curve(negexp(x, a = 1, b = 2), add = TRUE, col = "blue", lwd = 2)
legend("topright", lwd = 2, bty = "n", col = c("black", "red", "blue"),
       legend = c("b = 0.5", "b = 1", "b = 2"))
```

All start in the same place. `b` sets how fast it falls.

::: {.notes}
Contrast with the previous slide deliberately. a moves the curve up and down, b
changes its shape. Scale parameter versus rate parameter is the most transferable
idea in the chapter and shows up in every distribution in chapter 4.

Ask: what are the units of b? If x is in days, b is per day. That is what makes
it a rate, and why a mechanistic parameter beats a fitted constant.
:::

## Where does it fall by half?

For $f(x) = a\,e^{-b x}$ the book gives

$$x_{1/2} = \frac{\ln 2}{b}$$

In R, $\ln$ is `log()`. Check it:

```{r}
log(2) / 0.5
negexp(log(2) / 0.5, a = 1, b = 0.5)
```

This is a half-life. Bigger `b`, shorter half-life.

::: {.notes}
Still on the negative exponential, while its parameters are fresh. The point is
the habit: the book hands you a formula, you check it in one line rather than
trusting it.

Point at ln becoming log(). Anyone who types log10(2), a reasonable instinct after
Lecture 5, gets 0.602 instead of 1.386 and the check on the next line fails. That
is a good failure to have in the room.

Half-life is worth naming because they meet it again interpreting rate parameters
all course, and it is the same quantity as radioactive decay and drug clearance.
:::

## Try it against the data

::: {.type-it}
```{r}
#| fig-height: 3.2
plot(tadpoles$TBL, tadpoles$Kill, xlim = c(0, 40), ylim = c(0, 6),
     xlab = "Tadpole body length (mm)", ylab = "Number killed")
curve(negexp(x, a = 6, b = 0.1), add = TRUE, col = "red", lwd = 2)
```
:::

Falls back: **yes**. Rises first: **no**. It starts at its highest point.

::: {.notes}
The verdict, against the list: right ending, wrong start. It says the smallest
tadpoles are the most likely to be killed, and the data say 12 mm is the worst
size. So does the biology from the shape slide. No choice of a and b fixes it: a
negative exponential is always highest at x = 0.

One thing to be ready for, because a sharp student may find it once sum of squares
comes up: by that measure this curve fits almost as well as the best Ricker (about
27 against 25). That is not a contradiction. Sum of squares measures distance, not
whether the shape makes sense, and with 15 points, most of them zeros, both curves
fit imperfectly. We are choosing on shape and biology, which is the point of the
shape slide.
:::

## Put them together

```{r}
#| echo: false
#| fig-height: 3.6
curve(1 * x, from = 0, to = 8, ylim = c(0, 3), xlab = "x", ylab = "y",
      lty = 2, col = "grey40", lwd = 2)
curve(exp(-x),     add = TRUE, lty = 2, col = "grey40", lwd = 2)
curve(x * exp(-x), add = TRUE, col = "red", lwd = 3)
legend("topright", bty = "n", lwd = c(2, 2, 3), lty = c(2, 2, 1),
       col = c("grey40", "grey40", "red"),
       legend = c("x  (rises)", "exp(-x)  (falls)", "x * exp(-x)"))
```

The Holling curve got the **rise**. The exponential got the **fall**.
Multiply one that rises by one that falls: $x \times e^{-x}$ humps.

::: {.notes}
This is why the two failures came first: each one got half of the shape.
Multiplying them keeps the good half of each. Near zero the rising piece wins; far
out the exponential wins.

Be honest about one detail: the rising piece here is the simplest one, a straight
line through zero, not the Holling curve itself. A Holling curve times an
exponential would also hump, and building curves by combining pieces like that is
exactly what next lecture's bestiary is about.

Note 1 * x rather than x, for the curve() reason from earlier.
:::

## The Ricker, in math and in R {.smaller}

That product, with a parameter on each piece, is the **Ricker**: the standard
choice for humped, right-skewed ecological patterns.

:::: {.columns}

::: {.column width="50%"}
**Math**

$$f(x) = a\,x\,e^{-b x}$$

- $x$: the input, e.g. tadpole size
- $a$: initial slope, how steeply it rises from zero
- $b$: sets where it peaks, at $x = 1/b$ ($b > 0$)
:::

::: {.column width="50%"}
**R**

```{r}
ricker <- function(x, a = 1, b = 1){
  a * x * exp(-b * x)
}
```

- $a\,x$ is the rising piece
- $e^{-b x}$ is the falling piece
:::

::::

::: {.notes}
Not a new idea, just a name for the combination from the last slide. The a x is
the line that rises, and the e^(-bx) is the negative exponential from earlier, so
b means the same thing it did there: how fast the falling piece falls.

The parameter meanings are Bolker's: a is the initial slope and the peak sits at
1/b. The peak needs b > 0; with b at or below zero the falling piece does not fall,
so the curve never turns down. Both get checked in R shortly.
:::

## Your turn: type in the Ricker {background-color="#f0f4f8"}

Type this into your script and run it. You will use it for the rest of today.

::: {.type-it}
```{r}
#| eval: false
ricker <- function(x, a = 1, b = 1){
  a * x * exp(-b * x)
}

ricker(x = 2)          # should give 0.2706706
```
:::

Getting `could not find function "ricker"`? The definition has not been run yet.

::: {.notes}
Two minutes. Typing it, rather than copying, is the point: it is their first
function with exp() in it, and the check value tells them straight away whether
they got the brackets and the minus sign right.

Everything from here on uses ricker(), including both remaining exercises, so do
not skip this one even if time is short.
:::

## What happens at the ends?

You can just ask.

::: {.type-it}
```{r}
ricker(0, a = 1, b = 1)
ricker(c(10, 100, 1000), a = 1, b = 1)
```
:::

Starts at zero. Returns to zero. With a hump in between, that is every item on
our list.

. . .

Is starting at zero sensible for tadpoles?

::: {.notes}
Bolker takes limits analytically in 3.3.1. Plugging in numbers gets the same
answer and is available to everyone in the room.

Check it off against the shape slide: rises, peaks, falls back. This is the first
candidate that can do all three.

The last question is the biology. A Ricker says predation goes to zero for very
large tadpoles, which is plausible, and zero at size zero, which is a modeling
assumption you should notice you are making.
:::

## When you want the numbers

```{r}
xvec <- seq(from = 0, to = 8, length.out = 100)
yvec <- ricker(xvec, a = 1, b = 1)
head(yvec)
```

That worked because the function is **vectorized**.

. . .

If it is not, `sapply()` runs it one at a time:

```{r}
#| eval: false
yvec <- sapply(X = xvec, FUN = ricker)
```

::: {.notes}
Sets up the next slide, which finds the peak by computing lots of values.

This is the promise from Lecture 6 coming due. We said there were two better ways
than copy-pasting four blocks: a for loop, met in Lecture 7, and something
tidier. This is it.

A for loop makes you create the output vector first and fill it in. sapply()
collects the answers for you. Same idea, less bookkeeping.

The sapply half is the part to cut if you run long.
:::

## Where is the peak?

```{r}
xvec <- seq(0, 20, length.out = 10000)
yvec <- ricker(xvec, a = 1, b = 0.25)
xvec[which.max(yvec)]
```

. . .

The book does the calculus and gets $x_{\text{peak}} = \dfrac{1}{b}$.

```{r}
1 / 0.25
```

::: {.notes}
We care about the peak because the tadpole data have one, near 12 mm. If we know
where a Ricker peaks, we can put its peak where the data's peak is.
:::

## The peak depends only on `b`

```{r}
#| echo: false
curve(ricker(x, a = 1, b = 0.5), from = 0, to = 12, ylim = c(0, 1.6),
      xlab = "x", ylab = "y", lwd = 2)
curve(ricker(x, a = 1, b = 0.25), add = TRUE, col = "red",  lwd = 2)
curve(ricker(x, a = 1, b = 1),    add = TRUE, col = "blue", lwd = 2)
abline(v = c(2, 4, 1), lty = 3, col = c("black", "red", "blue"))
legend("topright", lwd = 2, bty = "n", col = c("red", "black", "blue"),
       legend = c("b = 0.25, peak at 4", "b = 0.5, peak at 2", "b = 1, peak at 1"))
```

Changing `a` stretches it vertically. The peak stays put.

::: {.notes}
This is what makes the two parameters worth separating, and what makes the
eyeball fit later possible: b places the peak, a sets its height.
:::

## Your turn {background-color="#f0f4f8"}

::: {.type-it}
```{r}
#| eval: false
# uses the ricker() you typed in earlier
a_try <- 1
b_try <- 0.1

curve(ricker(x, a = a_try, b = b_try), from = 0, to = 40)

# Find the peak numerically
xvec <- seq(0, 40, length.out = 10000)
yvec <- ricker(xvec, a = a_try, b = b_try)

xvec[which.max(yvec)]

1 / b_try            # does this agree?
```
:::

Now set `a_try <- 5`. Did the peak move?

What `b_try` puts the peak at **x = 20**?

::: {.notes}
Third pause, about 8 minutes, and the rehearsal for the eyeball fit.

Putting a_try and b_try at the top means the rest of the block is re-runnable
without editing anything inside the function calls. Same trick they will use on
the tadpole data.

The last question is the one that matters: they have to invert 1/b, which is
exactly the step they need in a moment on real data.
:::

## Back to the tadpoles

```{r}
#| echo: false
plot(tadpoles$TBL, tadpoles$Kill,
     xlab = "Tadpole body length (mm)", ylab = "Number killed",
     pch = 16, cex = 1.4, xlim = c(0, 40), ylim = c(0, 6))
```

Peak near **12 mm**. Height about **4** killed.

That is two facts. The Ricker has two parameters.

::: {.notes}
We chose the Ricker because it is the first candidate with the right shape. Now
put it where the data are.

Say the last line slowly. Reading two numbers off a picture determines two
parameters, which is why knowing what parameters mean matters: you get a starting
point without any fitting at all.
:::

## Eyeballing it

Peak is at `1/b`, and we want it at 12:

```{r}
b_eye <- 1 / 12
```

Height at the peak is $\dfrac{a}{b\,e}$, and we want 4. In R, $e$ is `exp(1)`:

```{r}
a_eye <- 4 * b_eye * exp(1)
a_eye
```

::: {.notes}
The height formula comes from substituting x = 1/b back into the Ricker, which
the book does. Quote it rather than deriving it; the useful skill is knowing such
a formula exists and using it.
:::

## How good is a fit?

$$\text{SS} = \sum_i \big(y_i - f(x_i)\big)^2$$

- $y_i$: each observed value, `tadpoles$Kill`
- $f(x_i)$: what the curve predicts there, `ricker(tadpoles$TBL, a, b)`
- $\sum_i$: add them all up, `sum()`

```{r}
sum((tadpoles$Kill - ricker(tadpoles$TBL, a = a_eye, b = b_eye))^2)
```

Smaller is better.

::: {.notes}
The first equation with a subscript, so read it aloud piece by piece and point at
the matching part of the R line each time. The i just means "for each data point".
Once they see that the sum is sum() and y_i is a whole column, the line of R reads
straight off the math.

No theory needed yet: total squared distance between data and curve, squared so
misses above and below do not cancel. Chapter 6 makes it respectable.
:::

## Your turn {background-color="#f0f4f8" .smaller}

::: {.type-it}
```{r}
#| eval: false
# uses the ricker() you typed in earlier
library(emdbook)
data(ReedfrogSizepred)
tadpoles <- ReedfrogSizepred

a_try <- 0.906                 # change these two
b_try <- 0.0833

plot(tadpoles$TBL, tadpoles$Kill, xlim = c(0, 40), ylim = c(0, 6))

curve(ricker(x, a = a_try, b = b_try), add = TRUE, col = "red", lwd = 2)

sum((tadpoles$Kill - ricker(tadpoles$TBL, a_try, b_try))^2)
```
:::

Starting score: **87.5**. Lower is better. Can you beat it?

::: {.notes}
Fourth pause, about 10 minutes. Then have them try to beat 87.5 by tweaking a and
b, which is the real exercise and is genuinely fun for about five minutes.

For reference, the best a Ricker can do on these data is about 24.9, at a = 1.01
and b = 0.151. That puts the peak at 6.6 mm rather than 12, which is itself a hint
that the Ricker is not quite the right shape.

They need library(emdbook) and data(ReedfrogSizepred) up somewhere while they
work.
:::

## It does not fit. Good.

```{r}
#| echo: false
plot(tadpoles$TBL, tadpoles$Kill,
     xlab = "Tadpole body length (mm)", ylab = "Number killed",
     pch = 16, cex = 1.4, xlim = c(0, 40), ylim = c(0, 6))
curve(ricker(x, a = 4 * (1/12) * exp(1), b = 1/12),
      add = TRUE, col = "red", lwd = 3)
```

Right kind of shape, but the right tail is too heavy. So you modify the function.

::: {.notes}
Same move as earlier, one level up. The Holling curve and the exponential each
failed in a way that told us what to try next. The Ricker has the right kind of
shape but fails in a smaller way: it falls too slowly, predicting kills at 30 mm
where the data say zero.

That is the book's own framing. Figure 3.1 shows the Ricker fitting badly and two
modified versions, the power-Ricker and a modified logistic, fitting well. Noticing
how a fit fails tells you what to change, and changing standard functions is next
lecture.
:::

## Before next class

Read Bolker **3.4 to 3.5**

Next time: the bestiary of functions, and modifying them to fit

Also: complete **3.6** (the R supplement)
